---
title: "miRNA Biomarkers in Cardiovascular Disease: NT-proBNP Comparative Analysis"
subtitle: "Supplementary Analysis for miRNAs in Cardiovascular Disease"
author: "BestAgeing Study Consortium"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    echo: false
    #code-fold: true
    #code-tools: true
    #code-summary: "Show the code"
    theme: cosmo
    fig-width: 10
    fig-height: 8
    df-print: kable
  pdf:
    toc: true
    number-sections: false
    colorlinks: true
    echo: false
    fig-width: 7
    fig-height: 5
    df-print: kable
    geometry:  # Set margins for PDF output default geometry: margin=1in (25.4mm all sides)
      - top=20mm
      - left=15mm
      - right=15mm
      - bottom=20mm
    mainfont: Times New Roman
    documentclass: report 
    classoption: [11pt] # Font size
    keep-tex: false # Keep the intermediate .tex file for debugging
bibliography: ../references.bib
---

# Executive Summary

This analysis investigates the diagnostic performance of individual miRNAs and machine learning-derived miRNA signatures compared to NT-proBNP in discriminating between dilated cardiomyopathy (DCM), heart failure with reduced ejection fraction (HFrEF), and healthy controls. We evaluate both univariate and multivariate diagnostic models, including incremental benefit analysis and clinical reclassification potential.

## Key Findings

- **Comparative Performance**: Top-performing individual miRNAs achieve diagnostic accuracy comparable to NT-proBNP
- **ML Signature Superiority**: Machine learning-derived miRNA signatures demonstrate enhanced diagnostic performance compared to individual miRNAs
- **Incremental Benefit**: Both individual miRNAs and ML signatures provide additional diagnostic information when combined with NT-proBNP
- **Clinical Utility**: Individual miRNAs and ML signatures demonstrate potential for patient reclassification in the NT-proBNP gray zone (125-450 pg/mL)

# Methods

## Study Population

The analysis includes patients from the BestAgeing study with:

- **DCM patients**: Dilated cardiomyopathy cases with available miRNA and NT-proBNP data
- **HFrEF patients**: Heart failure with reduced ejection fraction cases  
- **Controls**: Healthy individuals with NT-proBNP <450 pg/mL

## Data Processing

### NT-proBNP Preprocessing
- Controls with NT-proBNP >450 pg/mL excluded to ensure healthy control status
- Missing NT-proBNP values in controls imputed using control group mean
- All analyses performed on patients with complete covariate data

### miRNA Data
- Preprocessed expression values from established pipelines
- Standard quality control and normalization applied
- Missing values handled per established protocols

## Statistical Analysis

### Diagnostic Performance Assessment
1. **Univariate AUC**: ROC analysis for individual biomarkers
2. **Multivariate Models**: Logistic regression adjusting for age and sex
3. **Incremental Benefit**: AUC improvement when combining miRNAs with NT-proBNP

### Clinical Reclassification Analysis
- **Gray Zone Definition**: NT-proBNP 125-450 pg/mL (intermediate diagnostic confidence)
- **Reclassification Thresholds**: High-risk >0.7 probability, Low-risk <0.3 probability
- **Net Reclassification Improvement (NRI)**: Quantifying clinical utility

```{r setup}
#| include: false
#| code-fold: false

# Load required libraries
library(tidyverse)
library(limma)
library(edgeR)
library(ggplot2)
library(readxl)
library(writexl)
library(pROC)
library(janitor)
library(glue)
library(ggthemes)
library(workflows)
library(tune)
library(tidymodels)
library(colino)
library(knitr)
library(kableExtra)

# Set paths
data_path_bestageing2022 <- "/mnt/nas185/reich/rockerprojects/bestageing2022"
data_path_BestAgeing <- "/mnt/nas185/reich/BestAgeing"

# Load custom theme
source(file = glue("{data_path_bestageing2022}/scripts/helper/custom_ggplot_theme.R"))

# Create output directories if they don't exist
if (!dir.exists("output")) dir.create("output")
if (!dir.exists("figures")) dir.create("figures")
```

```{r helper-functions}
#| code-fold: true

# Helper functions
convert_mir_name <- function(name) {
  name <- gsub("mir", "miR", name)
  name <- gsub("_", "-", name)
  return(name)
}
convert_mir_name_V <- Vectorize(convert_mir_name)

# Data cleaning function for NT-proBNP analysis
clean_ntprobnp_data <- function(data, disease_name) {
  cat("Initial data dimensions:", dim(data), "\n")

  # Remove controls with NT-proBNP > 450
  controls_high_bnp <- data$disease == "control" &
    !is.na(data$ntprobnp) &
    data$ntprobnp > 450
  n_removed_controls <- sum(controls_high_bnp, na.rm = TRUE)

  if (n_removed_controls > 0) {
    data <- data[!controls_high_bnp, ]
    cat("Removed", n_removed_controls, "controls with NT-proBNP > 450\n")
  }

  # remove duplicates 2025-08-10
  data <- data[!duplicated(data$pat_id), ]

  # Mean imputation for controls with missing NT-proBNP
  control_ntprobnp_mean <- mean(data$ntprobnp[data$disease == "control"], na.rm = TRUE)
  missing_control_bnp <- data$disease == "control" & is.na(data$ntprobnp)
  n_imputed <- sum(missing_control_bnp, na.rm = TRUE)
  data$ntprobnp[missing_control_bnp] <- control_ntprobnp_mean

  if (n_imputed > 0) {
    cat(
      "Imputed NT-proBNP for", n_imputed, "controls using mean =",
      round(control_ntprobnp_mean, 2), "\n"
    )
  }

  # Store imputation value for filtering
  attr(data, "control_imputation_value") <- control_ntprobnp_mean

  # Remove patients with missing NT-proBNP for GLM analysis
  data_glm <- data[!is.na(data$ntprobnp), ]
  n_removed_missing <- nrow(data) - nrow(data_glm)

  cat("Final data dimensions for analysis:", dim(data_glm), "\n")

  return(list(full_data = data, glm_data = data_glm))
}
```

# Data Loading and Preprocessing

```{r}
#| label: load-ml-model-results

# created in ml_signature_ntprobnp_analysis.qmd

dcm_ml_result <- readRDS(glue("{data_path_bestageing2022}/revision2025/data/dcm_ml_result.rds"))
hfref_ml_result <- readRDS(glue("{data_path_bestageing2022}/revision2025/data/hfref_ml_result.rds"))

# Extract ML predictions for integration
dcm_ml_predictions <- dcm_ml_result$data %>%
  select(pat_id, .pred_dcm_logistic_reg_norm_selected_dcm) %>%
  rename(ml_signature_dcm = .pred_dcm_logistic_reg_norm_selected_dcm)

hfref_ml_predictions <- hfref_ml_result$data %>%
  select(pat_id, .pred_hfref_logistic_reg_norm_selected_hfref) %>%
  rename(ml_signature_hfref = .pred_hfref_logistic_reg_norm_selected_hfref)
```

```{r}
#| label: data-loading
#| include: false

# Load processed miRNA data
path2dataprocessed_dcm <- glue("{data_path_bestageing2022}/data/Rdata/processed_disease_data/001c_dcm_data01.rds")
path2dataprocessed_hfref <- glue("{data_path_bestageing2022}/data/Rdata/processed_disease_data/001c_hfref_data01.rds")

data_processed_dcm <- readRDS(path2dataprocessed_dcm)
data_processed_hfref <- readRDS(path2dataprocessed_hfref)

# Load clinical metadata
meta_dcm <- read_excel("/mnt/nas185/reich/BestAgeing/data/pheno_dcm.xlsx") %>%
  mutate(
    BestAgeingCode,
    phenodat = "dcm",
    ntprobnp = as.numeric(`DCM-Lab-ProBNP`),
    crea = as.numeric(`DCM-Lab-Cr`),
    lvdil_henry = as.numeric(`DCM-ECH-LVDil-Henry`),
    lvdil_lvedd = as.numeric(`DCM-ECH-LVDil-LVEDD`),
    lvsyst_ef = as.numeric(`DCM-ECH-LVSyst-EF`),
    symptoms = `DCM-PheDef-Symptoms`,
    wbc = as.numeric(`DCM-Lab-WBC`),
    .keep = "none"
  )

meta_hfref <- read_excel("/mnt/nas185/reich/BestAgeing/data/pheno_hfref.xlsx") %>%
  mutate(
    BestAgeingCode,
    phenodat = "hfref",
    ntprobnp = as.numeric(`SD-Lab-ProBNP`),
    crea = as.numeric(`SD-Lab-Cr`),
    lvdil_henry = as.numeric(`SD-ECH-LVDil-Henry`),
    lvdil_lvedd = as.numeric(`SD-ECH-LVDil-LVEDD`),
    lvsyst_ef = as.numeric(`SD-ECH-LVSyst-EF`),
    dyspnea = `SD-CliExam-Dyspnea`,
    wbc = as.numeric(`SD-Lab-WBC`),
    .keep = "none"
  )

meta_control <- read_excel("/mnt/nas185/reich/BestAgeing/data/pheno_controls.xlsx") %>%
  mutate(
    BestAgeingCode,
    phenodat = "control",
    ntprobnp = as.numeric(`Controls-Lab-ProBNP`),
    crea = as.numeric(`Controls-Lab-Cr`),
    .keep = "none"
  )

meta_all <- bind_rows(meta_dcm, meta_hfref, meta_control)

# Merge with processed data
dcm_merged <- data_processed_dcm %>%
  left_join(meta_all, by = c("pat_id" = "BestAgeingCode")) %>%
  mutate(preprocessing = "dcm_pipeline")

hfref_merged <- data_processed_hfref %>%
  left_join(meta_all, by = c("pat_id" = "BestAgeingCode")) %>%
  mutate(preprocessing = "hfref_pipeline")

# Clean data for NT-proBNP analysis
dcm_cleaned <- clean_ntprobnp_data(dcm_merged, "dcm")
hfref_cleaned <- clean_ntprobnp_data(hfref_merged, "hfref")

# Merge ML predictions with cleaned datasets
dcm_cleaned$full_data <- dcm_cleaned$full_data %>%
  left_join(dcm_ml_predictions, by = "pat_id")
dcm_cleaned$glm_data <- dcm_cleaned$glm_data %>%
  left_join(dcm_ml_predictions, by = "pat_id")

hfref_cleaned$full_data <- hfref_cleaned$full_data %>%
  left_join(hfref_ml_predictions, by = "pat_id")
hfref_cleaned$glm_data <- hfref_cleaned$glm_data %>%
  left_join(hfref_ml_predictions, by = "pat_id")
```



```{r}
#| label: sample-characteristics
#| echo: false

# Calculate sample characteristics
dcm_summary <- dcm_cleaned$glm_data %>%
  # remove imputed control values
  mutate(
    ntprobnp = ifelse(ntprobnp == attr(dcm_cleaned$glm_data, "control_imputation_value"), NA, ntprobnp),
  ) |>
  group_by(disease) %>%
  summarise(
    n = n(),
    age_mean = round(mean(age, na.rm = TRUE), 1),
    age_sd = round(sd(age, na.rm = TRUE), 1),
    female_pct = round(100 * mean(sex == "Female", na.rm = TRUE), 1),
    ntprobnp_median = round(median(ntprobnp, na.rm = TRUE), 0),
    ntprobnp_iqr_low = round(quantile(ntprobnp, 0.25, na.rm = TRUE), 0),
    ntprobnp_iqr_high = round(quantile(ntprobnp, 0.75, na.rm = TRUE), 0),
    .groups = "drop"
  ) %>%
  mutate(
    age_summary = paste0(age_mean, " ± ", age_sd),
    ntprobnp_summary = paste0(ntprobnp_median, " [", ntprobnp_iqr_low, "-", ntprobnp_iqr_high, "]"),
    cohort = "DCM"
  )

hfref_summary <- hfref_cleaned$glm_data %>%
  # remove imputed control values
  mutate(
    ntprobnp = ifelse(ntprobnp == attr(hfref_cleaned$glm_data, "control_imputation_value"), NA, ntprobnp),
  ) |>
  group_by(disease) %>%
  summarise(
    n = n(),
    age_mean = round(mean(age, na.rm = TRUE), 1),
    age_sd = round(sd(age, na.rm = TRUE), 1),
    female_pct = round(100 * mean(sex == "Female", na.rm = TRUE), 1),
    ntprobnp_median = round(median(ntprobnp, na.rm = TRUE), 0),
    ntprobnp_iqr_low = round(quantile(ntprobnp, 0.25, na.rm = TRUE), 0),
    ntprobnp_iqr_high = round(quantile(ntprobnp, 0.75, na.rm = TRUE), 0),
    .groups = "drop"
  ) %>%
  mutate(
    age_summary = paste0(age_mean, " ± ", age_sd),
    ntprobnp_summary = paste0(ntprobnp_median, " [", ntprobnp_iqr_low, "-", ntprobnp_iqr_high, "]"),
    cohort = "HFrEF"
  )

# Combine summaries
combined_summary <- bind_rows(dcm_summary, hfref_summary) %>%
  select(cohort, disease, n, age_summary, female_pct, ntprobnp_summary) %>%
  rename(
    "Cohort" = cohort,
    "Group" = disease,
    "N" = n,
    "Age (years)" = age_summary,
    "Female (%)" = female_pct,
    "NT-proBNP (pg/mL)" = ntprobnp_summary
  )

kable(combined_summary,
  caption = "Baseline characteristics of study populations",
  format = "html"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top")
```

# Differential Expression Analysis

This analysis evaluates miRNA diagnostic performance using multiple complementary approaches:

1. **Univariate Analysis**: Individual miRNA discrimination ability  
2. **Multivariate Models**: Adjusted for demographic confounders
3. **Incremental Benefit**: Added value when combined with NT-proBNP

```{r de-analysis-function}
#| include: false

run_de_analysis <- function(data_full, data_glm, disease_name) {
  data <- data_full

  # Filter to miRNA columns only
  all_filtered_mirnas <- data %>%
    select(-any_of(c(
      "disease", "age", "sex", "phenodat", "ntprobnp",
      "crea", "preprocessing"
    ))) %>%
    select(-contains("lvdil")) %>%
    select(-contains("lvsyst")) %>%
    select(-contains("dyspnea")) %>%
    select(-contains("symptoms")) %>%
    select(-contains("wbc"))

  n_mirnas <- ncol(all_filtered_mirnas) - 1 # Subtract pat_id column

  # Initialize result vectors
  aucs <- aucs_lowerci <- aucs_upperci <- auc_glm <- auc_glm_ntprobnp <- rep(NA, n_mirnas)
  pval_glm <- pval_glm_ntprobnp <- log2fc <- rep(NA, n_mirnas)
  name_mir <- rep(NA, n_mirnas)

  # NT-proBNP baseline model
  data_glm_complete <- data_glm[!is.na(data_glm$ntprobnp) & !is.na(data_glm$age) & !is.na(data_glm$sex), ]
  f_ntprobnp_only <- as.formula("disease ~ ntprobnp + age + sex")
  glm_ntprobnp_only <- tryCatch(
    {
      glm(f_ntprobnp_only, data = data_glm_complete, family = binomial())
    },
    error = function(e) NULL
  )

  auc_ntprobnp_only <- NA
  if (!is.null(glm_ntprobnp_only)) {
    pred_ntprobnp_only <- predict(glm_ntprobnp_only, type = "response")
    auc_ntprobnp_only <- as.numeric(roc(data_glm_complete$disease, pred_ntprobnp_only, quiet = TRUE)$auc)
  }

  # Analysis loop
  cont_index <- data$disease == "control"
  case_index <- data$disease == disease_name

  pb <- txtProgressBar(min = 0, max = n_mirnas, style = 3)

  for (i in 1:n_mirnas) {
    mirna_col <- i + 1 # Skip pat_id column
    cont <- as.numeric(data[cont_index, mirna_col, drop = TRUE])
    case <- as.numeric(data[case_index, mirna_col, drop = TRUE])

    # Remove NAs
    cont <- cont[!is.na(cont)]
    case <- case[!is.na(case)]

    if (length(cont) < 2 || length(case) < 2) {
      name_mir[i] <- names(data)[mirna_col]
      next
    }

    # Basic stats
    log2fc[i] <- mean(case) - mean(cont)
    name_mir[i] <- names(data)[mirna_col]

    # Univariate AUC with CI
    roc_obj <- roc(controls = cont, cases = case, quiet = TRUE)
    aucs[i] <- as.numeric(roc_obj$auc)
    auc_ci <- ci(roc_obj, quiet = TRUE)
    aucs_lowerci[i] <- auc_ci[1]
    aucs_upperci[i] <- auc_ci[3]

    # GLM models
    mirna_name <- names(data)[mirna_col]

    if (!mirna_name %in% names(data_glm)) next

    # Model 1: miRNA + age + sex
    data_complete <- data_glm[!is.na(data_glm[[mirna_name]]) &
      !is.na(data_glm$age) &
      !is.na(data_glm$sex), ]

    if (nrow(data_complete) > 10) {
      f1 <- as.formula(paste("disease ~", mirna_name, "+ age + sex"))
      glm1 <- tryCatch(
        {
          glm(f1, data = data_complete, family = binomial())
        },
        error = function(e) NULL
      )

      if (!is.null(glm1)) {
        pred1 <- predict(glm1, type = "response")
        auc_glm[i] <- as.numeric(roc(data_complete$disease, pred1, quiet = TRUE)$auc)
        pval_glm[i] <- coef(summary(glm1))[2, 4]
      }

      # Model 2: miRNA + age + sex + ntprobnp
      data_complete_nt <- data_complete[!is.na(data_complete$ntprobnp), ]

      if (nrow(data_complete_nt) > 10) {
        f2 <- as.formula(paste("disease ~", mirna_name, "+ age + sex + ntprobnp"))
        glm2 <- tryCatch(
          {
            glm(f2, data = data_complete_nt, family = binomial())
          },
          error = function(e) NULL
        )

        if (!is.null(glm2)) {
          pred2 <- predict(glm2, type = "response")
          auc_glm_ntprobnp[i] <- as.numeric(roc(data_complete_nt$disease, pred2, quiet = TRUE)$auc)
          pval_glm_ntprobnp[i] <- coef(summary(glm2))[2, 4]
        }
      }
    }

    setTxtProgressBar(pb, i)
  }
  close(pb)

  # Compile results
  results <- tibble(
    miRNA = name_mir,
    log2FoldChange = log2fc,
    auc_univariate = aucs,
    auc_lowerci = aucs_lowerci,
    auc_upperci = aucs_upperci,
    auc_glm = auc_glm,
    auc_glm_ntprobnp = auc_glm_ntprobnp,
    pval_glm = pval_glm,
    pval_glm_ntprobnp = pval_glm_ntprobnp,
    auc_ntprobnp_baseline = auc_ntprobnp_only
  ) %>%
    mutate(
      auc_univariate = ifelse(auc_univariate < 0.5, 1 - auc_univariate, auc_univariate),
      padj_glm = p.adjust(pval_glm, method = "holm"),
      padj_glm_ntprobnp = p.adjust(pval_glm_ntprobnp, method = "holm"),
      incremental_benefit_vs_baseline = auc_glm - auc_ntprobnp_baseline,
      incremental_benefit_combined = auc_glm_ntprobnp - auc_ntprobnp_baseline,
      incremental_benefit_miRNA_added = auc_glm_ntprobnp - auc_glm
    ) %>%
    arrange(desc(auc_univariate))

  return(results)
}
```

```{r run-analysis}
#| include: false

# Run DE analysis for both datasets
cat("Running DE analysis for DCM...\n")
dcm_results <- run_de_analysis(dcm_cleaned$full_data, dcm_cleaned$glm_data, "dcm")

cat("Running DE analysis for HFrEF...\n")
hfref_results <- run_de_analysis(hfref_cleaned$full_data, hfref_cleaned$glm_data, "hfref")

# Function to calculate ML signature performance
calculate_ml_signature_performance <- function(data_full, data_glm, disease_name, ml_signature_col) {
  # Use same structure as individual miRNA analysis
  data_glm_complete <- data_glm[!is.na(data_glm[[ml_signature_col]]) &
    !is.na(data_glm$ntprobnp) &
    !is.na(data_glm$age) &
    !is.na(data_glm$sex), ]

  # Univariate analysis
  cont <- data_full[data_full$disease == "control" & !is.na(data_full[[ml_signature_col]]), ml_signature_col, drop = TRUE]
  case <- data_full[data_full$disease == disease_name & !is.na(data_full[[ml_signature_col]]), ml_signature_col, drop = TRUE]

  if (length(cont) < 2 || length(case) < 2) {
    return(NULL)
  }

  # Univariate AUC with CI
  roc_obj <- roc(controls = cont, cases = case, quiet = TRUE)
  auc_univariate <- as.numeric(roc_obj$auc)
  auc_ci <- ci(roc_obj, quiet = TRUE)

  # NT-proBNP baseline model
  f_ntprobnp_only <- as.formula("disease ~ ntprobnp + age + sex")
  glm_ntprobnp_only <- tryCatch(
    glm(f_ntprobnp_only, data = data_glm_complete, family = binomial()),
    error = function(e) NULL
  )

  auc_ntprobnp_baseline <- NA
  if (!is.null(glm_ntprobnp_only)) {
    pred_ntprobnp_only <- predict(glm_ntprobnp_only, type = "response")
    auc_ntprobnp_baseline <- as.numeric(roc(data_glm_complete$disease, pred_ntprobnp_only, quiet = TRUE)$auc)
  }

  # Model 1: ML signature + age + sex
  f1 <- as.formula(paste("disease ~", ml_signature_col, "+ age + sex"))
  glm1 <- tryCatch(
    glm(f1, data = data_glm_complete, family = binomial()),
    error = function(e) NULL
  )

  auc_glm <- pval_glm <- NA
  if (!is.null(glm1)) {
    pred1 <- predict(glm1, type = "response")
    auc_glm <- as.numeric(roc(data_glm_complete$disease, pred1, quiet = TRUE)$auc)
    pval_glm <- coef(summary(glm1))[2, 4]
  }

  # Model 2: ML signature + age + sex + ntprobnp
  f2 <- as.formula(paste("disease ~", ml_signature_col, "+ age + sex + ntprobnp"))
  glm2 <- tryCatch(
    glm(f2, data = data_glm_complete, family = binomial()),
    error = function(e) NULL
  )

  auc_glm_ntprobnp <- pval_glm_ntprobnp <- NA
  if (!is.null(glm2)) {
    pred2 <- predict(glm2, type = "response")
    auc_glm_ntprobnp <- as.numeric(roc(data_glm_complete$disease, pred2, quiet = TRUE)$auc)
    pval_glm_ntprobnp <- coef(summary(glm2))[2, 4]
  }

  # Return results in same format as miRNA results
  return(tibble(
    miRNA = "miRNA Signature (ML)",
    log2FoldChange = mean(case) - mean(cont),
    auc_univariate = auc_univariate,
    auc_lowerci = auc_ci[1],
    auc_upperci = auc_ci[3],
    auc_glm = auc_glm,
    auc_glm_ntprobnp = auc_glm_ntprobnp,
    pval_glm = pval_glm,
    pval_glm_ntprobnp = pval_glm_ntprobnp,
    auc_ntprobnp_baseline = auc_ntprobnp_baseline,
    padj_glm = pval_glm, # No multiple testing correction for single comparison
    padj_glm_ntprobnp = pval_glm_ntprobnp,
    incremental_benefit_vs_baseline = auc_glm - auc_ntprobnp_baseline,
    incremental_benefit_combined = auc_glm_ntprobnp - auc_ntprobnp_baseline,
    incremental_benefit_miRNA_added = auc_glm_ntprobnp - auc_glm
  ))
}

# Calculate ML signature performance
cat("Calculating ML signature performance...\n")
dcm_ml_performance <- calculate_ml_signature_performance(
  dcm_cleaned$full_data, dcm_cleaned$glm_data, "dcm", "ml_signature_dcm"
)

hfref_ml_performance <- calculate_ml_signature_performance(
  hfref_cleaned$full_data, hfref_cleaned$glm_data, "hfref", "ml_signature_hfref"
)

# Calculate NT-proBNP baseline performance
ntprobnp_performance <- function(data, disease_name) {
  if (all(is.na(data$ntprobnp))) {
    return(NA)
  }

  data_clean <- data %>% filter(!is.na(ntprobnp))
  cont <- data_clean[data_clean$disease == "control", "ntprobnp", drop = TRUE]
  case <- data_clean[data_clean$disease == disease_name, "ntprobnp", drop = TRUE]

  roc_obj <- roc(controls = cont, cases = case, quiet = TRUE)
  return(as.numeric(roc_obj$auc))
}

ntprobnp_auc_dcm <- ntprobnp_performance(dcm_cleaned$full_data, "dcm")
ntprobnp_auc_hfref <- ntprobnp_performance(hfref_cleaned$full_data, "hfref")
```

```{r}
#| label: ntprobnp-performance

cat("NT-proBNP univariate AUC DCM:", round(ntprobnp_auc_dcm, 3), "\n")
cat("NT-proBNP univariate AUC HFrEF:", round(ntprobnp_auc_hfref, 3), "\n")
```

# Results

## Baseline Biomarker Performance

```{r baseline-performance}
#| echo: false

# NT-proBNP multivariate baseline from results
ntprobnp_baseline_dcm <- unique(dcm_results$auc_ntprobnp_baseline)[1]
ntprobnp_baseline_hfref <- unique(hfref_results$auc_ntprobnp_baseline)[1]

baseline_performance <- tibble(
  Cohort = c("DCM", "DCM", "HFrEF", "HFrEF"),
  Model = c(
    "NT-proBNP (univariate)", "NT-proBNP + age + sex",
    "NT-proBNP (univariate)", "NT-proBNP + age + sex"
  ),
  AUC = c(
    ntprobnp_auc_dcm, ntprobnp_baseline_dcm,
    ntprobnp_auc_hfref, ntprobnp_baseline_hfref
  )
) %>%
  mutate(AUC = round(AUC, 3))

kable(baseline_performance,
  caption = "NT-proBNP diagnostic performance across cohorts",
  format = "html"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top")
```

NT-proBNP demonstrates excellent diagnostic performance in both cohorts, with multivariate models (adjusted for age and sex) achieving AUC values of `r round(ntprobnp_baseline_dcm, 3)` (DCM) and `r round(ntprobnp_baseline_hfref, 3)` (HFrEF).

## Top-Performing miRNAs

### DCM Cohort

```{r}
#| label: dcm-top-results
#| echo: false

# Combine individual miRNAs with ML signature
dcm_combined_results <- dcm_results %>%
  filter(!is.na(auc_univariate) & miRNA != "disease") %>%
  head(9) # Take top 9 individual miRNAs

if (!is.null(dcm_ml_performance)) {
  dcm_combined_results <- bind_rows(dcm_combined_results, dcm_ml_performance)
}

dcm_top <- dcm_combined_results %>%
  select(
    miRNA, auc_univariate, auc_glm, auc_glm_ntprobnp,
    incremental_benefit_vs_baseline, incremental_benefit_combined,
    padj_glm
  ) %>%
  arrange(desc(auc_univariate)) %>%
  mutate(
    miRNA = ifelse(miRNA == "miRNA Signature (ML)", miRNA, convert_mir_name_V(miRNA)),
    across(where(is.numeric), ~ round(.x, 3))
  )

kable(dcm_top,
  col.names = c(
    "miRNA", "Univariate AUC", "Multivariate AUC",
    "Combined AUC", "vs Baseline Δ", "Combined Δ", "Adj. P-value"
  ),
  caption = "Top 10 miRNAs in DCM cohort with incremental benefit analysis",
  format = "html"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(
    " " = 1, "Individual Performance" = 2,
    "Combined Models" = 2, "Incremental Benefit" = 2
  ))
```

### HFrEF Cohort

```{r hfref-top-results}
#| echo: false

# Combine individual miRNAs with ML signature
hfref_combined_results <- hfref_results %>%
  filter(!is.na(auc_univariate) & miRNA != "disease") %>%
  head(9) # Take top 9 individual miRNAs

if (!is.null(hfref_ml_performance)) {
  hfref_combined_results <- bind_rows(hfref_combined_results, hfref_ml_performance)
}

hfref_top <- hfref_combined_results %>%
  select(
    miRNA, auc_univariate, auc_glm, auc_glm_ntprobnp,
    incremental_benefit_vs_baseline, incremental_benefit_combined,
    padj_glm
  ) %>%
  arrange(desc(auc_univariate)) %>%
  mutate(
    miRNA = ifelse(miRNA == "miRNA Signature (ML)", miRNA, convert_mir_name_V(miRNA)),
    across(where(is.numeric), ~ round(.x, 3))
  )

kable(hfref_top,
  col.names = c(
    "miRNA", "Univariate AUC", "Multivariate AUC",
    "Combined AUC", "vs Baseline Δ", "Combined Δ", "Adj. P-value"
  ),
  caption = "Top 10 miRNAs in HFrEF cohort with incremental benefit analysis",
  format = "html"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(
    " " = 1, "Individual Performance" = 2,
    "Combined Models" = 2, "Incremental Benefit" = 2
  ))
```

## Performance Visualization

```{r}
#| label: performance-visualization
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

# Prepare data for visualization - include ML signature
dcm_viz_data <- dcm_results %>%
  filter(!is.na(auc_univariate) & miRNA != "disease") %>%
  slice_max(auc_univariate, n = 14) # Take top 14 individual miRNAs

if (!is.null(dcm_ml_performance)) {
  dcm_viz_data <- bind_rows(dcm_viz_data, dcm_ml_performance)
}

dcm_viz <- dcm_viz_data %>%
  mutate(
    miRNA_clean = ifelse(miRNA == "miRNA Signature (ML)", miRNA, convert_mir_name_V(miRNA)),
    miRNA_clean = fct_reorder(miRNA_clean, auc_univariate),
    cohort = "DCM"
  )

hfref_viz_data <- hfref_results %>%
  filter(!is.na(auc_univariate) & miRNA != "disease") %>%
  slice_max(auc_univariate, n = 14) # Take top 14 individual miRNAs

if (!is.null(hfref_ml_performance)) {
  hfref_viz_data <- bind_rows(hfref_viz_data, hfref_ml_performance)
}

hfref_viz <- hfref_viz_data %>%
  mutate(
    miRNA_clean = ifelse(miRNA == "miRNA Signature (ML)", miRNA, convert_mir_name_V(miRNA)),
    miRNA_clean = fct_reorder(miRNA_clean, auc_univariate),
    cohort = "HFrEF"
  )

combined_viz <- bind_rows(dcm_viz, hfref_viz)

# AUC comparison plot
p1 <- combined_viz %>%
  ggplot(aes(x = miRNA_clean, y = auc_univariate, fill = cohort)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(
    yintercept = c(ntprobnp_auc_dcm, ntprobnp_auc_hfref),
    linetype = "dashed", alpha = 0.7, size = 1
  ) +
  coord_flip() +
  scale_fill_manual(values = c("DCM" = "#1f77b4", "HFrEF" = "#ff7f0e")) +
  labs(
    title = "Top miRNA Diagnostic Performance vs NT-proBNP",
    subtitle = "Dashed lines represent NT-proBNP univariate AUC for each cohort",
    x = "miRNA",
    y = "AUC (95% CI)",
    fill = "Cohort"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )

# Incremental benefit plot
p2 <- combined_viz %>%
  filter(!is.na(incremental_benefit_combined)) %>%
  ggplot(aes(x = miRNA_clean, y = incremental_benefit_combined, fill = cohort)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c("DCM" = "#1f77b4", "HFrEF" = "#ff7f0e")) +
  labs(
    title = "Incremental Benefit: miRNA + NT-proBNP vs NT-proBNP Alone",
    subtitle = "Positive values indicate improved performance when combining biomarkers",
    x = "miRNA",
    y = "ΔAUC (Combined vs Baseline)",
    fill = "Cohort"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )

print(p1)
print(p2)
```

## Clinical Interpretation

### Key Performance Metrics

- **Best miRNA Performance**: 
  - DCM: `r convert_mir_name(dcm_top$miRNA[1])` (AUC = `r dcm_top$auc_univariate[1]`)
  - HFrEF: `r convert_mir_name(hfref_top$miRNA[1])` (AUC = `r hfref_top$auc_univariate[1]`)

- **Incremental Benefit**: Several miRNAs provide meaningful diagnostic improvement when combined with NT-proBNP, with ΔAUC values up to `r max(c(dcm_top$incremental_benefit_combined, hfref_top$incremental_benefit_combined), na.rm = TRUE)`

### Clinical Utility Assessment

The analysis demonstrates that:

1. **Comparable Individual Performance**: Top miRNAs achieve diagnostic accuracy approaching that of NT-proBNP
2. **Additive Information**: Combining miRNAs with NT-proBNP improves diagnostic performance beyond either biomarker alone
3. **Disease-Specific Patterns**: Different miRNAs show optimal performance in DCM versus HFrEF, suggesting distinct pathophysiological mechanisms

# Gray Zone Analysis

```{r}
#| label: gray-zone-analysis

gray_zone_analysis <- function(data, results, disease_name, limit_disyplay = 5) {
  # Use data with actual NT-proBNP measurements only
  data_original <- data %>% filter(!is.na(ntprobnp))

  # Remove imputed control values
  imputation_value <- attr(data, "control_imputation_value")
  if (!is.null(imputation_value)) {
    data_original <- data_original %>%
      filter(!(disease == "control" & abs(ntprobnp - imputation_value) < 0.01))
  }

  if (nrow(data_original) < 20) {
    return(tibble())
  }

  # Calculate quartiles for disease group
  data_disease_only <- data_original %>% filter(disease == disease_name)
  median_ntprobnp <- median(data_disease_only$ntprobnp, na.rm = TRUE)

  # Gray zone (lower than median)
  gray_zone_data <- data_original %>% filter(ntprobnp < median_ntprobnp)

  # Test top miRNAs in gray zone
  top_mirnas <- results %>%
    filter(miRNA != "disease", !is.na(auc_univariate)) %>%
    slice_max(auc_univariate, n = 5)

  gray_zone_results <- tibble()

  for (i in 1:min(limit_disyplay, nrow(top_mirnas))) { # Limit to top mirnas for display
    mir_name <- top_mirnas$miRNA[i]

    if (!mir_name %in% names(gray_zone_data)) next

    gray_glm_data <- gray_zone_data %>%
      filter(!is.na(.data[[mir_name]]), !is.na(age), !is.na(sex)) %>%
      mutate(disease_binary = ifelse(disease == disease_name, 1, 0))

    if (nrow(gray_glm_data) < 15) next

    # Models in gray zone
    f1 <- as.formula(paste("disease_binary ~", mir_name, "+ age + sex"))
    f2 <- as.formula("disease_binary ~ ntprobnp + age + sex")
    f3 <- as.formula(paste("disease_binary ~", mir_name, "+ ntprobnp + age + sex"))

    glm1 <- tryCatch(glm(f1, data = gray_glm_data, family = binomial()), error = function(e) NULL)
    glm2 <- tryCatch(glm(f2, data = gray_glm_data, family = binomial()), error = function(e) NULL)
    glm3 <- tryCatch(glm(f3, data = gray_glm_data, family = binomial()), error = function(e) NULL)

    if (!is.null(glm1) && !is.null(glm2) && !is.null(glm3)) {
      auc1 <- as.numeric(roc(gray_glm_data$disease_binary, predict(glm1, type = "response"), quiet = TRUE)$auc)
      auc2 <- as.numeric(roc(gray_glm_data$disease_binary, predict(glm2, type = "response"), quiet = TRUE)$auc)
      auc3 <- as.numeric(roc(gray_glm_data$disease_binary, predict(glm3, type = "response"), quiet = TRUE)$auc)

      # Statistical test for incremental benefit
      stat_test_p <- anova(glm2, glm3, test = "Chisq")$`Pr(>Chi)`[2]

      # Console output for incremental analysis
      cat(sprintf("  %s Gray Zone Incremental Analysis:\n", mir_name))
      cat(sprintf("    miRNA + age + sex:           AUC=%.3f\n", auc1))
      cat(sprintf("    NT-proBNP + age + sex:       AUC=%.3f\n", auc2))
      cat(sprintf("    Combined + age + sex:        AUC=%.3f\n", auc3))
      cat(sprintf("    Incremental benefit:         Δ=%.3f (p=%.3f)\n", auc3 - auc2, stat_test_p))

      gray_zone_results <- bind_rows(
        gray_zone_results,
        tibble(
          miRNA = mir_name,
          auc_mirna_model = auc1,
          auc_ntprobnp_model = auc2,
          auc_combined_model = auc3,
          incremental_benefit = auc3 - auc2,
          statistical_test_p = stat_test_p,
          n_gray_zone = nrow(gray_glm_data),
          n_cases_gray = sum(gray_glm_data$disease_binary == 1),
          n_controls_gray = sum(gray_glm_data$disease_binary == 0),
          median_ntprobnp_threshold = median_ntprobnp
        )
      )
    }
  }

  return(gray_zone_results)
}


# Run gray zone analysis
dcm_gray <- gray_zone_analysis(dcm_cleaned$full_data, dcm_results, "dcm")
hfref_gray <- gray_zone_analysis(hfref_cleaned$full_data, hfref_results, "hfref")

# Add ML signature to gray zone analysis
gray_zone_analysis_ml <- function(data, disease_name, ml_signature_col, ml_name) {
  # Use data with actual NT-proBNP measurements only
  data_original <- data %>% filter(!is.na(ntprobnp))

  # Remove imputed control values
  imputation_value <- attr(data, "control_imputation_value")
  if (!is.null(imputation_value)) {
    data_original <- data_original %>%
      filter(!(disease == "control" & abs(ntprobnp - imputation_value) < 0.01))
  }

  if (nrow(data_original) < 20) {
    return(tibble())
  }

  # Calculate median for disease group
  data_disease_only <- data_original %>% filter(disease == disease_name)
  median_ntprobnp <- median(data_disease_only$ntprobnp, na.rm = TRUE)

  # Gray zone (lower than median)
  gray_zone_data <- data_original %>% filter(ntprobnp < median_ntprobnp)

  # Test ML signature in gray zone
  if (!ml_signature_col %in% names(gray_zone_data)) {
    return(tibble())
  }

  gray_glm_data <- gray_zone_data %>%
    filter(!is.na(.data[[ml_signature_col]]), !is.na(age), !is.na(sex)) %>%
    mutate(disease_binary = ifelse(disease == disease_name, 1, 0))

  if (nrow(gray_glm_data) < 15) {
    return(tibble())
  }

  # Models in gray zone
  f1 <- as.formula(paste("disease_binary ~", ml_signature_col, "+ age + sex"))
  f2 <- as.formula("disease_binary ~ ntprobnp + age + sex")
  f3 <- as.formula(paste("disease_binary ~", ml_signature_col, "+ ntprobnp + age + sex"))

  glm1 <- tryCatch(glm(f1, data = gray_glm_data, family = binomial()), error = function(e) NULL)
  glm2 <- tryCatch(glm(f2, data = gray_glm_data, family = binomial()), error = function(e) NULL)
  glm3 <- tryCatch(glm(f3, data = gray_glm_data, family = binomial()), error = function(e) NULL)

  if (!is.null(glm1) && !is.null(glm2) && !is.null(glm3)) {
    auc1 <- as.numeric(roc(gray_glm_data$disease_binary, predict(glm1, type = "response"), quiet = TRUE)$auc)
    auc2 <- as.numeric(roc(gray_glm_data$disease_binary, predict(glm2, type = "response"), quiet = TRUE)$auc)
    auc3 <- as.numeric(roc(gray_glm_data$disease_binary, predict(glm3, type = "response"), quiet = TRUE)$auc)

    # Statistical test for incremental benefit
    stat_test_p <- anova(glm2, glm3, test = "Chisq")$`Pr(>Chi)`[2]

    # Console output for incremental analysis
    cat(sprintf("  %s Gray Zone Incremental Analysis:\n", ml_name))
    cat(sprintf("    ML signature + age + sex:       AUC=%.3f\n", auc1))
    cat(sprintf("    NT-proBNP + age + sex:           AUC=%.3f\n", auc2))
    cat(sprintf("    Combined + age + sex:            AUC=%.3f\n", auc3))
    cat(sprintf("    Incremental benefit:             Δ=%.3f (p=%.3f)\n", auc3 - auc2, stat_test_p))

    return(tibble(
      miRNA = ml_name,
      auc_mirna_model = auc1,
      auc_ntprobnp_model = auc2,
      auc_combined_model = auc3,
      incremental_benefit = auc3 - auc2,
      statistical_test_p = stat_test_p,
      n_gray_zone = nrow(gray_glm_data),
      n_cases_gray = sum(gray_glm_data$disease_binary == 1),
      n_controls_gray = sum(gray_glm_data$disease_binary == 0),
      median_ntprobnp_threshold = median_ntprobnp
    ))
  }

  return(tibble())
}

# Add ML signature to gray zone results
if (!is.null(dcm_ml_performance)) {
  dcm_ml_gray <- gray_zone_analysis_ml(dcm_cleaned$full_data, "dcm", "ml_signature_dcm", "miRNA Signature (ML)")
  if (nrow(dcm_ml_gray) > 0) {
    dcm_gray <- bind_rows(dcm_gray, dcm_ml_gray)
  }
}

if (!is.null(hfref_ml_performance)) {
  hfref_ml_gray <- gray_zone_analysis_ml(hfref_cleaned$full_data, "hfref", "ml_signature_hfref", "miRNA Signature (ML)")
  if (nrow(hfref_ml_gray) > 0) {
    hfref_gray <- bind_rows(hfref_gray, hfref_ml_gray)
  }
}
```

```{r}
#| label: roc-plot-fn

# Plot ROC curves for gray zone analysis
plot_gray_zone_roc <- function(data, results, disease_name, mirna_names) {
  # Use data with actual NT-proBNP measurements only
  data_original <- data %>% filter(!is.na(ntprobnp))

  # Remove imputed control values
  imputation_value <- attr(data, "control_imputation_value")
  if (!is.null(imputation_value)) {
    data_original <- data_original %>%
      filter(!(disease == "control" & abs(ntprobnp - imputation_value) < 0.01))
  }

  # Calculate median for disease group
  data_disease_only <- data_original %>% filter(disease == disease_name)
  median_ntprobnp <- median(data_disease_only$ntprobnp, na.rm = TRUE)

  # Gray zone data
  gray_zone_data <- data_original %>% filter(ntprobnp < median_ntprobnp)

  # Handle single miRNA vs vector
  if (is.character(mirna_names) && length(mirna_names) == 1) {
    mirna_names <- mirna_names
  }

  # Create plots for each miRNA
  plot_list <- list()

  for (mir_name in mirna_names) {
    if (!mir_name %in% names(gray_zone_data)) next

    gray_glm_data <- gray_zone_data %>%
      filter(!is.na(.data[[mir_name]]), !is.na(age), !is.na(sex)) %>%
      mutate(disease_binary = ifelse(disease == disease_name, 1, 0))

    if (nrow(gray_glm_data) < 15) next

    # Models
    f2 <- as.formula("disease_binary ~ ntprobnp + age + sex")
    f3 <- as.formula(paste("disease_binary ~", mir_name, "+ ntprobnp + age + sex"))

    glm2 <- tryCatch(glm(f2, data = gray_glm_data, family = binomial()), error = function(e) NULL)
    glm3 <- tryCatch(glm(f3, data = gray_glm_data, family = binomial()), error = function(e) NULL)

    if (!is.null(glm2) && !is.null(glm3)) {
      # Calculate ROCs
      roc2 <- roc(gray_glm_data$disease_binary, predict(glm2, type = "response"), quiet = TRUE)
      roc3 <- roc(gray_glm_data$disease_binary, predict(glm3, type = "response"), quiet = TRUE)

      auc2 <- as.numeric(roc2$auc)
      auc3 <- as.numeric(roc3$auc)
      stat_test_p <- anova(glm2, glm3, test = "Chisq")$`Pr(>Chi)`[2]

      # Create ROC data
      roc_data <- bind_rows(
        tibble(
          specificity = roc2$specificities,
          sensitivity = roc2$sensitivities,
          model = "NT-proBNP + covariates",
          auc = auc2
        ),
        tibble(
          specificity = roc3$specificities,
          sensitivity = roc3$sensitivities,
          model = "Combined + covariates",
          auc = auc3
        )
      )

      # Create plot
      p <- ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
        geom_line(size = 1.2) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
        scale_color_manual(values = c("NT-proBNP + covariates" = "#d62728", "Combined + covariates" = "#1f77b4")) +
        labs(
          title = convert_mir_name_V(mir_name), # paste("Gray Zone ROC Curves:", convert_mir_name_V(mir_name)),
          subtitle = element_blank(), # paste(disease_name, "patients with NT-proBNP < median (n =", nrow(gray_glm_data), ")"),
          x = "1 - Specificity",
          y = "Sensitivity",
          color = "Model"
        ) +
        theme_minimal(base_size = 14) +
        theme(
          legend.position = "none",
          panel.grid.minor = element_blank()
        ) +
        coord_equal() +
        annotate("text",
          x = 0.6, y = 0.4,
          label = paste0("NT-proBNP AUC: ", round(auc2, 3)),
          hjust = 0, size = 3.5, color = "#d62728"
        ) +
        annotate("text",
          x = 0.6, y = 0.35,
          label = paste0("Combined AUC: ", round(auc3, 3)),
          hjust = 0, size = 3.5, color = "#1f77b4"
        ) +
        annotate("text",
          x = 0.6, y = 0.25,
          label = paste0(
            "Δ AUC: ", round(auc3 - auc2, 3),
            "\np = ", ifelse(stat_test_p < 0.001, "<0.001", round(stat_test_p, 3)), ""
          ),
          hjust = 0, size = 3.5, color = "black"
        )

      plot_list[[mir_name]] <- p
    }
  }

  # Return single plot or combined plots
  if (length(plot_list) == 1) {
    return(plot_list[[1]])
  } else if (length(plot_list) > 1) {
    # Combine plots using patchwork
    library(patchwork)
    return(wrap_plots(plot_list, ncol = 2))
  }

  return(NULL)
}
```

```{r}
#| echo: false
#| label: gray-zone-roc-plots
#| fig-width: 10
#| fig-height: 10

# Function to plot ROC curves including ML signature
plot_gray_zone_roc_with_ml <- function(data, results, disease_name, mirna_names, ml_col = NULL, ml_name = NULL) {
  plots <- list()

  # Plot regular miRNAs
  regular_mirnas <- mirna_names[mirna_names != "miRNA Signature (ML)"]
  if (length(regular_mirnas) > 0) {
    regular_plot <- plot_gray_zone_roc(data, results, disease_name, regular_mirnas)
    if (!is.null(regular_plot)) plots <- append(plots, list(regular_plot))
  }

  # Plot ML signature if requested
  if ("miRNA Signature (ML)" %in% mirna_names && !is.null(ml_col)) {
    ml_plot <- plot_gray_zone_roc_ml(data, disease_name, ml_col, ml_name)
    if (!is.null(ml_plot)) plots <- append(plots, list(ml_plot))
  }

  # Return combined plots
  if (length(plots) == 1) {
    return(plots[[1]])
  } else if (length(plots) > 1) {
    library(patchwork)
    return(wrap_plots(plots, ncol = 2))
  }

  return(NULL)
}

# Function to plot ROC for ML signature
plot_gray_zone_roc_ml <- function(data, disease_name, ml_col, ml_name) {
  # Use data with actual NT-proBNP measurements only
  data_original <- data %>% filter(!is.na(ntprobnp))

  # Remove imputed control values
  imputation_value <- attr(data, "control_imputation_value")
  if (!is.null(imputation_value)) {
    data_original <- data_original %>%
      filter(!(disease == "control" & abs(ntprobnp - imputation_value) < 0.01))
  }

  # Calculate median for disease group
  data_disease_only <- data_original %>% filter(disease == disease_name)
  median_ntprobnp <- median(data_disease_only$ntprobnp, na.rm = TRUE)

  # Gray zone data
  gray_zone_data <- data_original %>% filter(ntprobnp < median_ntprobnp)

  if (!ml_col %in% names(gray_zone_data)) {
    return(NULL)
  }

  gray_glm_data <- gray_zone_data %>%
    filter(!is.na(.data[[ml_col]]), !is.na(age), !is.na(sex)) %>%
    mutate(disease_binary = ifelse(disease == disease_name, 1, 0))

  if (nrow(gray_glm_data) < 15) {
    return(NULL)
  }

  # Models
  f2 <- as.formula("disease_binary ~ ntprobnp + age + sex")
  f3 <- as.formula(paste("disease_binary ~", ml_col, "+ ntprobnp + age + sex"))

  glm2 <- tryCatch(glm(f2, data = gray_glm_data, family = binomial()), error = function(e) NULL)
  glm3 <- tryCatch(glm(f3, data = gray_glm_data, family = binomial()), error = function(e) NULL)

  if (!is.null(glm2) && !is.null(glm3)) {
    # Calculate ROCs
    roc2 <- roc(gray_glm_data$disease_binary, predict(glm2, type = "response"), quiet = TRUE)
    roc3 <- roc(gray_glm_data$disease_binary, predict(glm3, type = "response"), quiet = TRUE)

    auc2 <- as.numeric(roc2$auc)
    auc3 <- as.numeric(roc3$auc)
    stat_test_p <- anova(glm2, glm3, test = "Chisq")$`Pr(>Chi)`[2]

    # Create ROC data
    roc_data <- bind_rows(
      tibble(
        specificity = roc2$specificities,
        sensitivity = roc2$sensitivities,
        model = "NT-proBNP + covariates",
        auc = auc2
      ),
      tibble(
        specificity = roc3$specificities,
        sensitivity = roc3$sensitivities,
        model = "Combined + covariates",
        auc = auc3
      )
    )

    # Create plot
    p <- ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
      geom_line(size = 1.2) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
      scale_color_manual(values = c("NT-proBNP + covariates" = "#d62728", "Combined + covariates" = "#1f77b4")) +
      labs(
        title = ml_name,
        x = "1 - Specificity",
        y = "Sensitivity",
        color = "Model"
      ) +
      theme_minimal(base_size = 14) +
      theme(
        legend.position = "none",
        panel.grid.minor = element_blank()
      ) +
      coord_equal() +
      annotate("text",
        x = 0.6, y = 0.4,
        label = paste0("NT-proBNP AUC: ", round(auc2, 3)),
        hjust = 0, size = 3.5, color = "#d62728"
      ) +
      annotate("text",
        x = 0.6, y = 0.35,
        label = paste0("Combined AUC: ", round(auc3, 3)),
        hjust = 0, size = 3.5, color = "#1f77b4"
      ) +
      annotate("text",
        x = 0.6, y = 0.25,
        label = paste0(
          "Δ AUC: ", round(auc3 - auc2, 3),
          "\np = ", ifelse(stat_test_p < 0.001, "<0.001", round(stat_test_p, 3))
        ),
        hjust = 0, size = 3.5, color = "black"
      )

    return(p)
  }

  return(NULL)
}

# Plot ROC curves for top miRNAs including ML signature
if (nrow(dcm_gray) > 0) {
  top_dcm_mirnas <- dcm_gray$miRNA[1:min(4, nrow(dcm_gray))]
  dcm_roc_plot <- plot_gray_zone_roc_with_ml(dcm_cleaned$full_data, dcm_results, "dcm", top_dcm_mirnas, "ml_signature_dcm", "miRNA Signature (ML)")
  print(dcm_roc_plot)
}

if (nrow(hfref_gray) > 0) {
  top_hfref_mirnas <- hfref_gray$miRNA[1:min(4, nrow(hfref_gray))]
  hfref_roc_plot <- plot_gray_zone_roc_with_ml(hfref_cleaned$full_data, hfref_results, "hfref", top_hfref_mirnas, "ml_signature_hfref", "miRNA Signature (ML)")
  print(hfref_roc_plot)
}
```

```{r}
#| echo: false
#| label: gray-zone-results

if (nrow(dcm_gray) > 0) {
  dcm_gray_display <- dcm_gray %>%
    mutate(
      miRNA = ifelse(miRNA == "miRNA Signature (ML)", miRNA, convert_mir_name_V(miRNA)),
      across(contains("auc"), ~ round(.x, 3)),
      incremental_benefit = round(incremental_benefit, 3),
      statistical_test_p = round(statistical_test_p, 3),
      median_ntprobnp_threshold = round(median_ntprobnp_threshold, 0)
    ) %>%
    select(
      miRNA, auc_mirna_model, auc_ntprobnp_model, auc_combined_model,
      incremental_benefit, statistical_test_p, n_gray_zone, median_ntprobnp_threshold
    )

  kable(dcm_gray_display,
    col.names = c(
      "miRNA", "miRNA AUC", "NT-proBNP AUC", "Combined AUC",
      "Incremental Δ", "p-value", "N (Gray Zone)", "Threshold (pg/mL)"
    ),
    caption = "DCM Gray Zone Analysis: Performance in patients with NT-proBNP below median",
    format = "html"
  ) %>%
    kable_styling(bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE) %>%
    add_header_above(c(" " = 1, "Model Performance" = 5, "Sample Size" = 2))
}

if (nrow(hfref_gray) > 0) {
  hfref_gray_display <- hfref_gray %>%
    mutate(
      miRNA = ifelse(miRNA == "miRNA Signature (ML)", miRNA, convert_mir_name_V(miRNA)),
      across(contains("auc"), ~ round(.x, 3)),
      incremental_benefit = round(incremental_benefit, 3),
      statistical_test_p = round(statistical_test_p, 3),
      median_ntprobnp_threshold = round(median_ntprobnp_threshold, 0)
    ) %>%
    select(
      miRNA, auc_mirna_model, auc_ntprobnp_model, auc_combined_model,
      incremental_benefit, statistical_test_p, n_gray_zone, median_ntprobnp_threshold
    )

  kable(hfref_gray_display,
    col.names = c(
      "miRNA", "miRNA AUC", "NT-proBNP AUC", "Combined AUC",
      "Incremental Δ", "p-value", "N (Gray Zone)", "Threshold (pg/mL)"
    ),
    caption = "HFrEF Gray Zone Analysis: Performance in patients with NT-proBNP below median",
    format = "html"
  ) %>%
    kable_styling(bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE) %>%
    add_header_above(c(" " = 1, "Model Performance" = 5, "Sample Size" = 2))
}
```

The gray zone analysis reveals that miRNAs provide particularly valuable diagnostic information in patients with NT-proBNP levels below the disease-specific median, where traditional biomarker performance may be suboptimal.


# Reclassification Analysis 

```{r}
#| label: reclassification-analysis-clinical-thresholds

reclassification_analysis <- function(data, results, disease_name) {
  # Use data with actual NT-proBNP measurements only
  data_original <- data %>%
    filter(!is.na(ntprobnp))

  # Remove imputed control values using the stored imputation value
  imputation_value <- attr(data, "control_imputation_value")
  if (!is.null(imputation_value)) {
    data_original <- data_original %>%
      filter(!(disease == "control" & abs(ntprobnp - imputation_value) < 0.01))
    cat(
      "Removed imputed control values (",
      round(imputation_value, 2),
      " pg/mL)\n"
    )
  }

  if (nrow(data_original) < 20) {
    cat(
      "Insufficient non-imputed data for reclassification analysis in",
      disease_name,
      "\n"
    )
    return(list())
  }

  # Clinical NT-proBNP thresholds
  rule_out_threshold <- 125 # Rule out HF if < 125 pg/mL
  rule_in_threshold <- 450 # Rule in HF if > 450 pg/mL
  # Gray zone: 125-450 pg/mL

  cat("\n=== CLINICAL RECLASSIFICATION ANALYSIS:", disease_name, "===\n")
  cat("NT-proBNP Clinical Thresholds:\n")
  cat("  Rule-out (<125 pg/mL): Low probability of HF\n")
  cat("  Gray zone (125-450 pg/mL): Intermediate probability\n")
  cat("  Rule-in (>450 pg/mL): High probability of HF\n")

  # Classify patients by NT-proBNP thresholds
  data_classified <- data_original %>%
    mutate(
      ntprobnp_class = case_when(
        ntprobnp < rule_out_threshold ~ "Rule-out (<125)",
        ntprobnp >= rule_out_threshold & ntprobnp <= rule_in_threshold ~ "Gray zone (125-450)",
        ntprobnp > rule_in_threshold ~ "Rule-in (>450)",
        TRUE ~ NA_character_
      ),
      is_case = ifelse(disease == disease_name, 1, 0)
    )

  # Summary by zones
  zone_summary <- data_classified %>%
    group_by(ntprobnp_class) %>%
    summarise(
      n_total = n(),
      n_cases = sum(is_case),
      n_controls = sum(1 - is_case),
      case_percentage = round(100 * n_cases / n_total, 1),
      .groups = "drop"
    )

  cat("\nPatient distribution by NT-proBNP zones:\n")
  print(zone_summary)

  # Get top miRNAs for reclassification analysis
  top_mirnas <- results %>%
    filter(miRNA != "disease", !is.na(auc_univariate)) %>%
    slice_max(auc_univariate, n = 3)

  reclassification_results <- list()

  for (i in 1:nrow(top_mirnas)) {
    mir_name <- top_mirnas$miRNA[i]

    if (!mir_name %in% names(data_classified)) {
      next
    }

    # Focus on gray zone patients (most clinically relevant)
    gray_zone_patients <- data_classified %>%
      filter(
        ntprobnp_class == "Gray zone (125-450)",
        !is.na(.data[[mir_name]]),
        !is.na(age),
        !is.na(sex)
      )

    if (nrow(gray_zone_patients) < 10) {
      next
    }

    # Build GLM model for reclassification
    f_model <- as.formula(paste("is_case ~", mir_name, "+ age + sex"))
    glm_model <- tryCatch(
      glm(f_model, data = gray_zone_patients, family = binomial()),
      error = function(e) NULL
    )

    if (!is.null(glm_model)) {
      # Get predicted probabilities
      gray_zone_patients$pred_prob <- predict(glm_model, type = "response")

      # Define reclassification thresholds based on predicted probability
      # High confidence: >0.7 probability
      # Low confidence: <0.3 probability
      # Intermediate: 0.3-0.7 probability

      gray_zone_reclassified <- gray_zone_patients %>%
        mutate(
          mirna_reclass = case_when(
            pred_prob > 0.7 ~ "High risk (miRNA)",
            pred_prob < 0.3 ~ "Low risk (miRNA)",
            TRUE ~ "Intermediate (miRNA)"
          ),
          # Create reclassification categories
          reclass_category = case_when(
            mirna_reclass == "High risk (miRNA)" & is_case == 1 ~ "Correct rule-in",
            mirna_reclass == "High risk (miRNA)" & is_case == 0 ~ "Incorrect rule-in",
            mirna_reclass == "Low risk (miRNA)" & is_case == 0 ~ "Correct rule-out",
            mirna_reclass == "Low risk (miRNA)" & is_case == 1 ~ "Incorrect rule-out",
            TRUE ~ "No reclassification"
          )
        )

      # Calculate reclassification metrics
      reclass_summary <- gray_zone_reclassified %>%
        count(reclass_category) %>%
        mutate(percentage = round(100 * n / nrow(gray_zone_reclassified), 1))

      # Net Reclassification Improvement (NRI)
      correct_reclass <- sum(reclass_summary$n[
        reclass_summary$reclass_category %in%
          c("Correct rule-in", "Correct rule-out")
      ])
      incorrect_reclass <- sum(reclass_summary$n[
        reclass_summary$reclass_category %in%
          c("Incorrect rule-in", "Incorrect rule-out")
      ])

      nri <- (correct_reclass - incorrect_reclass) /
        nrow(gray_zone_reclassified)

      cat(sprintf(
        "\n%s Reclassification in Gray Zone (n=%d):\n",
        mir_name,
        nrow(gray_zone_reclassified)
      ))
      print(reclass_summary)
      cat(sprintf("Net Reclassification Improvement: %.3f\n", nri))

      reclassification_results[[mir_name]] <- list(
        data = gray_zone_reclassified,
        summary = reclass_summary,
        nri = nri,
        zone_summary = zone_summary
      )
    }
  }

  return(reclassification_results)
}

# Run reclassification analysis
dcm_reclass <- reclassification_analysis(dcm_cleaned$full_data, dcm_results, "dcm")
hfref_reclass <- reclassification_analysis(hfref_cleaned$full_data, hfref_results, "hfref")

# Function for ML signature reclassification analysis
reclassification_analysis_ml <- function(data, disease_name, ml_col, ml_name) {
  # Use data with actual NT-proBNP measurements only
  data_original <- data %>%
    filter(!is.na(ntprobnp))

  # Remove imputed control values using the stored imputation value
  imputation_value <- attr(data, "control_imputation_value")
  if (!is.null(imputation_value)) {
    data_original <- data_original %>%
      filter(!(disease == "control" & abs(ntprobnp - imputation_value) < 0.01))
  }

  if (nrow(data_original) < 20) {
    return(list())
  }

  # Clinical NT-proBNP thresholds
  rule_out_threshold <- 125
  rule_in_threshold <- 450

  # Classify patients by NT-proBNP thresholds
  data_classified <- data_original %>%
    mutate(
      ntprobnp_class = case_when(
        ntprobnp < rule_out_threshold ~ "Rule-out (<125)",
        ntprobnp >= rule_out_threshold & ntprobnp <= rule_in_threshold ~ "Gray zone (125-450)",
        ntprobnp > rule_in_threshold ~ "Rule-in (>450)",
        TRUE ~ NA_character_
      ),
      is_case = ifelse(disease == disease_name, 1, 0)
    )

  # Focus on gray zone patients (most clinically relevant)
  gray_zone_patients <- data_classified %>%
    filter(
      ntprobnp_class == "Gray zone (125-450)",
      !is.na(.data[[ml_col]]),
      !is.na(age),
      !is.na(sex)
    )

  if (nrow(gray_zone_patients) < 10) {
    return(list())
  }

  # Build GLM model for reclassification
  f_model <- as.formula(paste("is_case ~", ml_col, "+ age + sex"))
  glm_model <- tryCatch(
    glm(f_model, data = gray_zone_patients, family = binomial()),
    error = function(e) NULL
  )

  if (!is.null(glm_model)) {
    # Get predicted probabilities
    gray_zone_patients$pred_prob <- predict(glm_model, type = "response")

    # Define reclassification thresholds based on predicted probability
    gray_zone_reclassified <- gray_zone_patients %>%
      mutate(
        mirna_reclass = case_when(
          pred_prob > 0.7 ~ "High risk (ML signature)",
          pred_prob < 0.3 ~ "Low risk (ML signature)",
          TRUE ~ "Intermediate (ML signature)"
        ),
        # Create reclassification categories
        reclass_category = case_when(
          mirna_reclass == "High risk (ML signature)" & is_case == 1 ~ "Correct rule-in",
          mirna_reclass == "High risk (ML signature)" & is_case == 0 ~ "Incorrect rule-in",
          mirna_reclass == "Low risk (ML signature)" & is_case == 0 ~ "Correct rule-out",
          mirna_reclass == "Low risk (ML signature)" & is_case == 1 ~ "Incorrect rule-out",
          TRUE ~ "No reclassification"
        )
      )

    # Calculate reclassification metrics
    reclass_summary <- gray_zone_reclassified %>%
      count(reclass_category) %>%
      mutate(percentage = round(100 * n / nrow(gray_zone_reclassified), 1))

    # Net Reclassification Improvement (NRI)
    correct_reclass <- sum(reclass_summary$n[
      reclass_summary$reclass_category %in%
        c("Correct rule-in", "Correct rule-out")
    ])
    incorrect_reclass <- sum(reclass_summary$n[
      reclass_summary$reclass_category %in%
        c("Incorrect rule-in", "Incorrect rule-out")
    ])

    nri <- (correct_reclass - incorrect_reclass) /
      nrow(gray_zone_reclassified)

    cat(sprintf(
      "\n%s Reclassification in Gray Zone (n=%d):\n",
      ml_name,
      nrow(gray_zone_reclassified)
    ))
    print(reclass_summary)
    cat(sprintf("Net Reclassification Improvement: %.3f\n", nri))

    return(list(
      data = gray_zone_reclassified,
      summary = reclass_summary,
      nri = nri
    ))
  }

  return(list())
}

# Add ML signature reclassification analysis
if (!is.null(dcm_ml_performance) && "ml_signature_dcm" %in% names(dcm_cleaned$full_data)) {
  dcm_ml_reclass <- reclassification_analysis_ml(dcm_cleaned$full_data, "dcm", "ml_signature_dcm", "miRNA Signature")
  if (length(dcm_ml_reclass) > 0) {
    dcm_reclass[["miRNA Signature"]] <- dcm_ml_reclass
  }
}

if (!is.null(hfref_ml_performance) && "ml_signature_hfref" %in% names(hfref_cleaned$full_data)) {
  hfref_ml_reclass <- reclassification_analysis_ml(hfref_cleaned$full_data, "hfref", "ml_signature_hfref", "miRNA Signature")
  if (length(hfref_ml_reclass) > 0) {
    hfref_reclass[["miRNA Signature"]] <- hfref_ml_reclass
  }
}
```

```{r}
#| label: reclassification_percentage_plot

plot_reclassification_percentage <- function(reclass_result, mirna_names, disease_name) {
  # Handle both single miRNA and vector of miRNAs
  if (is.character(mirna_names) && length(mirna_names) == 1) {
    mirna_names <- mirna_names
  }

  # Reorder to put miRNA Signature first if it exists
  if ("miRNA Signature" %in% mirna_names) {
    mirna_names <- c("miRNA Signature", mirna_names[mirna_names != "miRNA Signature"])
  }

  # Prepare data for all requested miRNAs
  plot_data_list <- list()

  for (mirna_name in mirna_names) {
    if (is.null(reclass_result[[mirna_name]])) {
      cat("No reclassification data for", mirna_name, "\n")
      next
    }

    summary_data <- reclass_result[[mirna_name]]$summary

    # Combine incorrect categories and calculate percentages
    mirna_data <- summary_data %>%
      mutate(
        category_clean = case_when(
          reclass_category == "Correct rule-in" ~ "Correct RI",
          reclass_category == "Correct rule-out" ~ "Correct RO",
          reclass_category == "No reclassification" ~ "No Reclassification",
          reclass_category %in% c("Incorrect rule-in", "Incorrect rule-out") ~ "Incorrect Reclassification"
        )
      ) %>%
      group_by(category_clean) %>%
      summarise(n = sum(n), .groups = "drop") %>%
      mutate(
        percentage = round(100 * n / sum(n), 1),
        mirna = ifelse(mirna_name == "miRNA Signature", "miRNA Signature", convert_mir_name_V(mirna_name)),
        total_n = sum(n)
      )

    plot_data_list[[mirna_name]] <- mirna_data
  }

  # Combine all data
  plot_data <- bind_rows(plot_data_list) %>%
    mutate(
      category_clean = factor(category_clean,
        levels = c("Incorrect Reclassification", "Correct RO", "No Reclassification", "Correct RI")
      ),
      mirna = factor(mirna, levels = ifelse(mirna_names == "miRNA Signature", "miRNA Signature", convert_mir_name_V(mirna_names)))
    )

  # Choose appropriate plot based on number of miRNAs
  if (length(mirna_names) == 1) {
    # Single miRNA - horizontal bars with category colors
    ggplot(plot_data, aes(x = percentage, y = category_clean, fill = category_clean)) +
      geom_col(width = 0.7) +
      geom_text(aes(label = paste0(percentage, "% (n=", n, ")")),
        hjust = -0.1, size = 3.5
      ) +
      scale_fill_manual(values = c(
        "Correct RI" = "#1f77b4", # Blue
        "Correct RO" = "#2ca02c", # Green
        "No Reclassification" = "#d3d3d3", # Light gray
        "Incorrect Reclassification" = "#ff7f0e" # Orange
      )) +
      scale_x_continuous(limits = c(0, max(plot_data$percentage) * 1.3)) +
      scale_y_continuous(
        limits = c(0, max(plot_data$percentage) * 1.15),
        breaks = seq(0, ceiling(max(plot_data$percentage) / 20) * 20, by = 20),
        labels = scales::label_percent(accuracy = 1, scale = 1)
      ) +
      labs(
        title = paste("Reclassification Impact:", unique(plot_data$mirna)),
        subtitle = paste(disease_name, "patients (n =", unique(plot_data$total_n), ")"),
        x = "Percentage of Patients",
        y = ""
      ) +
      theme_minimal(base_size = 16) +
      theme(
        legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()
      )
  } else {
    # Multiple miRNAs - dodged bars with miRNA colors matching ROC plot

    # Create color mapping consistent with ROC plot
    mirna_levels <- levels(plot_data$mirna)
    colors <- character()
    mirna_color_palette <- c("#33A02C", "#FF7F00", "#6A3D9A") # Same as ROC plot

    for (i in seq_along(mirna_levels)) {
      if (mirna_levels[i] == "miRNA Signature") {
        colors[mirna_levels[i]] <- "#1F78B4" # Dark blue for signature
      } else {
        # Assign colors to individual miRNAs
        mirna_idx <- sum(mirna_levels[1:(i - 1)] != "miRNA Signature") + 1
        colors[mirna_levels[i]] <- mirna_color_palette[((mirna_idx - 1) %% length(mirna_color_palette)) + 1]
      }
    }

    ggplot(plot_data, aes(x = category_clean, y = percentage, fill = mirna)) +
      geom_col(position = position_dodge(width = 0.8), width = 0.7) +
      geom_text(aes(label = paste0(percentage, "%")),
        position = position_dodge(width = 0.8),
        hjust = -0.1, size = 3, angle = 0
      ) +
      scale_y_continuous(
        limits = c(0, max(plot_data$percentage) * 1.15),
        breaks = seq(0, ceiling(max(plot_data$percentage) / 20) * 20, by = 20),
        labels = scales::label_percent(accuracy = 1, scale = 1)
      ) +
      scale_fill_manual(values = colors) +
      # wrap legend rows
      guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
      labs(
        title = "",
        subtitle = paste(disease_name, "patients"),
        x = "",
        y = "",
        fill = "miRNA"
      ) +
      theme_minimal(base_size = 16) +
      theme(
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.box = "horizontal",
        legend.text = element_text(size = 11),
        legend.key.width = unit(14, "pt"),
        legend.key.height = unit(12, "pt"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(t = 10, r = 30, b = 10, l = 10)
      ) +
      coord_flip(clip = "off")
  }
}

p1 <- plot_reclassification_percentage(dcm_reclass, names(dcm_reclass), "DCM")
p2 <- plot_reclassification_percentage(hfref_reclass, names(hfref_reclass), "ICM")
```

## Reclassification Impact Comparison Plot

```{r}
#| label: reclassification-plots-print1
#| fig-width: 8
#| fig-height: 5

p1
ggsave("revision2025/figures/reclassification_dcm_plot.png", p1, width = 8, height = 5, dpi = 300)
```

```{r}
#| label: reclassification-plots-print2
#| fig-width: 8
#| fig-height: 5

p2
ggsave("revision2025/figures/reclassification_icm_plot.png", p2, width = 8, height = 5, dpi = 300)
```

# Summary ROC Plots

```{r}
#| label: summary-roc-plots-fn

# Simple function to create ROC plot with top miRNAs + ML signature (Gray Zone)
create_summary_roc <- function(data_full, results, ml_performance, disease_name, ml_col, title, show_ci = FALSE) {
  # Use data with actual NT-proBNP measurements only
  data_original <- data_full %>% filter(!is.na(ntprobnp))

  # Remove imputed control values
  imputation_value <- attr(data_full, "control_imputation_value")
  if (!is.null(imputation_value)) {
    data_original <- data_original %>%
      filter(!(disease == "control" & abs(ntprobnp - imputation_value) < 0.01))
  }

  # Calculate median for disease group (Gray Zone = below median)
  data_disease_only <- data_original %>% filter(disease == disease_name)
  median_ntprobnp <- median(data_disease_only$ntprobnp, na.rm = TRUE)

  # Gray zone data
  data_roc <- data_original %>%
    filter(ntprobnp < median_ntprobnp, !is.na(age), !is.na(sex)) %>%
    mutate(disease_binary = ifelse(disease == disease_name, 1, 0))

  # Get top 3 miRNAs
  top_mirnas <- results %>%
    filter(!is.na(auc_univariate) & miRNA != "disease") %>%
    slice_max(auc_univariate, n = 3)

  roc_data_list <- list()

  # NT-proBNP baseline
  glm_nt <- glm(disease_binary ~ ntprobnp + age + sex, data = data_roc, family = binomial())
  roc_nt <- roc(data_roc$disease_binary, predict(glm_nt, type = "response"), quiet = TRUE)
  roc_data_list[["NT-proBNP + covariates"]] <- tibble(
    specificity = roc_nt$specificities,
    sensitivity = roc_nt$sensitivities,
    biomarker = paste0("NT-proBNP (AUC: ", sprintf("%.3f", as.numeric(roc_nt$auc)), ")")
  )

  # Store ROC objects for CI calculation
  roc_objects <- list()
  roc_objects[["NT-proBNP + covariates"]] <- roc_nt

  # Top 3 miRNAs (combined with NT-proBNP)
  for (i in 1:nrow(top_mirnas)) {
    mir_name <- top_mirnas$miRNA[i]
    if (mir_name %in% names(data_roc)) {
      data_complete <- data_roc[!is.na(data_roc[[mir_name]]), ]
      glm_combined <- glm(as.formula(paste("disease_binary ~", mir_name, "+ ntprobnp + age + sex")),
        data = data_complete, family = binomial()
      )
      roc_combined <- roc(data_complete$disease_binary, predict(glm_combined, type = "response"), quiet = TRUE)
      roc_data_list[[paste("Combined +", convert_mir_name_V(mir_name))]] <- tibble(
        specificity = roc_combined$specificities,
        sensitivity = roc_combined$sensitivities,
        biomarker = paste0(convert_mir_name_V(mir_name), " (AUC: ", sprintf("%.3f", as.numeric(roc_combined$auc)), ")")
      )
    }
  }

  # ML signature (combined with NT-proBNP)
  if (!is.null(ml_performance) && ml_col %in% names(data_roc)) {
    data_ml <- data_roc[!is.na(data_roc[[ml_col]]), ]
    glm_combined_ml <- glm(as.formula(paste("disease_binary ~", ml_col, "+ ntprobnp + age + sex")),
      data = data_ml, family = binomial()
    )
    roc_combined_ml <- roc(data_ml$disease_binary, predict(glm_combined_ml, type = "response"), quiet = TRUE)
    roc_data_list[["Combined + ML Signature"]] <- tibble(
      specificity = roc_combined_ml$specificities,
      sensitivity = roc_combined_ml$sensitivities,
      biomarker = paste0("miRNA Signature (AUC: ", sprintf("%.3f", as.numeric(roc_combined_ml$auc)), ")")
    )
    # Store ROC object for CI calculation
    roc_objects[["Combined + ML Signature"]] <- roc_combined_ml
  }

  combined_roc <- bind_rows(roc_data_list)

  # Create alpha and size mappings for highlighting
  combined_roc <- combined_roc %>%
    mutate(
      curve_type = case_when(
        grepl("NT-proBNP", biomarker) ~ "reference",
        grepl("miRNA Signature", biomarker) ~ "signature",
        TRUE ~ "individual_mirna"
      ),
      alpha_val = case_when(
        curve_type == "reference" ~ 1.0,
        curve_type == "signature" ~ 1.0,
        TRUE ~ 0.6
      ),
      size_val = case_when(
        curve_type == "reference" ~ 1.5,
        curve_type == "signature" ~ 1.5,
        TRUE ~ 1.0
      )
    )

  # Reorder biomarkers: signature first, then miRNAs, then NT-proBNP (reference) last
  biomarker_order <- c(
    combined_roc$biomarker[combined_roc$curve_type == "signature"],
    combined_roc$biomarker[combined_roc$curve_type == "individual_mirna"],
    combined_roc$biomarker[combined_roc$curve_type == "reference"]
  )
  biomarker_order <- unique(biomarker_order)
  combined_roc$biomarker <- factor(combined_roc$biomarker, levels = biomarker_order)

  # Colors: signature = dark blue, reference = red, miRNAs = other colors
  n_mirnas <- sum(combined_roc$curve_type == "individual_mirna")

  # Create color vector based on curve type, not order
  colors <- character()
  for (biomarker in biomarker_order) {
    curve_type <- unique(combined_roc$curve_type[combined_roc$biomarker == biomarker])
    if (curve_type == "signature") {
      colors[biomarker] <- "#1F78B4" # dark blue for signature
    } else if (curve_type == "reference") {
      colors[biomarker] <- "#E31A1C" # red for reference
    } else if (curve_type == "individual_mirna") {
      # Assign colors to miRNAs in order they appear
      mirna_colors <- c("#33A02C", "#FF7F00", "#6A3D9A")
      mirna_idx <- sum(names(colors) %in% combined_roc$biomarker[combined_roc$curve_type == "individual_mirna"]) + 1
      colors[biomarker] <- mirna_colors[((mirna_idx - 1) %% length(mirna_colors)) + 1]
    }
  }

  ggplot(combined_roc, aes(x = 1 - specificity, y = sensitivity, color = biomarker)) +
    geom_line(aes(alpha = I(alpha_val), size = I(size_val))) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
    scale_color_manual(values = colors) +
    labs(title = title, x = "1 - Specificity", y = "Sensitivity", color = "Biomarker") +
    theme_minimal(base_size = 16) +
    theme(
      legend.title = element_blank(),
      legend.position = "right"
    ) +
    coord_equal() +
    guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.2)))
}

# DCM Gray Zone ROC
roc_dcm <- create_summary_roc(dcm_cleaned$full_data, dcm_results, dcm_ml_performance,
  "dcm", "ml_signature_dcm",
  title = ""
)

# HFrEF Gray Zone ROC
roc_hfref <- create_summary_roc(hfref_cleaned$full_data, hfref_results, hfref_ml_performance,
  "hfref", "ml_signature_hfref",
  title = ""
)
```

```{r}
#| label: summary-roc-plots-dcm
#| fig-width: 10
#| fig-height: 6

roc_dcm
ggsave("revision2025/figures/roc_dcm_gray_zone.png", roc_dcm, width = 10, height = 6, dpi = 300)
```


```{r}
#| label: summary-roc-plots-hfref
#| fig-width: 10
#| fig-height: 6

roc_hfref
ggsave("revision2025/figures/roc_hfref_gray_zone.png", roc_hfref, width = 10, height = 6, dpi = 300)
```


# Conclusions

This analysis demonstrates that miRNAs represent promising complementary biomarkers to NT-proBNP in cardiovascular disease diagnosis. Key findings include:

1. **Diagnostic Equivalence**: Top individual miRNAs achieve diagnostic accuracy comparable to NT-proBNP in both DCM and HFrEF
2. **ML Signature Performance**: The machine learning-derived miRNA signature shows enhanced diagnostic performance compared to individual miRNAs, potentially offering superior clinical utility
3. **Incremental Benefit**: Both individual miRNAs and the ML signature provide consistent diagnostic improvement when combined with NT-proBNP
4. **Clinical Utility**: miRNAs and ML signatures show particular value in diagnostically challenging scenarios (gray zone patients)
5. **Reclassification Potential**: The ML signature demonstrates substantial patient reclassification capability in the NT-proBNP gray zone
6. **Translation Potential**: Performance characteristics support clinical development of both individual miRNA biomarkers and multi-miRNA signature assays

These results provide a strong foundation for advancing miRNA biomarkers toward clinical implementation in cardiovascular disease diagnosis, with the ML-derived multi-miRNA signature showing particularly promising performance characteristics as part of multi-biomarker diagnostic strategies.

```{r save-results}
#| include: false

# Save key results for downstream analyses
write_csv(dcm_results, "output/dcm_mirna_ntprobnp_results.csv")
write_csv(hfref_results, "output/hfref_mirna_ntprobnp_results.csv")

# Save ML signature performance
if (!is.null(dcm_ml_performance)) {
  write_csv(dcm_ml_performance, "output/dcm_ml_signature_performance.csv")
}
if (!is.null(hfref_ml_performance)) {
  write_csv(hfref_ml_performance, "output/hfref_ml_signature_performance.csv")
}

# Save gray zone analysis (including ML signatures)
if (nrow(dcm_gray) > 0) {
  write_csv(dcm_gray, "output/dcm_gray_zone_analysis.csv")
}
if (nrow(hfref_gray) > 0) {
  write_csv(hfref_gray, "output/hfref_gray_zone_analysis.csv")
}

# Save reclassification results (including ML signatures)
if (length(dcm_reclass) > 0) {
  # Create summary of all reclassification results
  dcm_reclass_summary <- map_dfr(names(dcm_reclass), function(mir_name) {
    if (!is.null(dcm_reclass[[mir_name]]$nri)) {
      tibble(
        miRNA = mir_name,
        nri = dcm_reclass[[mir_name]]$nri,
        n_gray_zone = nrow(dcm_reclass[[mir_name]]$data)
      )
    }
  })
  if (nrow(dcm_reclass_summary) > 0) {
    write_csv(dcm_reclass_summary, "output/dcm_reclassification_summary.csv")
  }
}

if (length(hfref_reclass) > 0) {
  # Create summary of all reclassification results
  hfref_reclass_summary <- map_dfr(names(hfref_reclass), function(mir_name) {
    if (!is.null(hfref_reclass[[mir_name]]$nri)) {
      tibble(
        miRNA = mir_name,
        nri = hfref_reclass[[mir_name]]$nri,
        n_gray_zone = nrow(hfref_reclass[[mir_name]]$data)
      )
    }
  })
  if (nrow(hfref_reclass_summary) > 0) {
    write_csv(hfref_reclass_summary, "output/hfref_reclassification_summary.csv")
  }
}

cat("Analysis complete. Results saved to output/ directory.\n")
cat("ML signature integration completed successfully.\n")
```

---

*Analysis conducted using R `r R.version.string` with miRNA expression data from the BestAgeing study. For questions regarding methodology or data access, please contact the study investigators.*