---
title: "ML Model Performance by Study Center"
subtitle: "Center-specific performance analysis"
author: "BestAgeing Study Consortium"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    echo: false
    theme: cosmo
    fig-width: 12
    fig-height: 8
    df-print: kable
---

```{r setup}
#| include: false

library(tidyverse)
library(pROC)
library(glue)
library(knitr)
library(kableExtra)
library(patchwork)

# Set paths
data_path_bestageing2022 <- "/mnt/nas185/reich/rockerprojects/bestageing2022"

# Create output directories
if (!dir.exists("output")) dir.create("output")
if (!dir.exists("figures")) dir.create("figures")
if (!dir.exists("scripts2025/figures/ml_center_performance")) {
  dir.create("scripts2025/figures/ml_center_performance", recursive = TRUE)
}
```

```{r load-data}
#| include: false

# Load ML results with predictions
dcm_ml_result <- readRDS("revision2025/data/dcm_ml_result_full.rds")
hfref_ml_result <- readRDS("revision2025/data/hfref_ml_result_full.rds")
acs_ml_result <- readRDS("revision2025/data/acs_ml_result.rds")
cad_ml_result <- readRDS("revision2025/data/cad_ml_result.rds")

# Extract data
dcm_data <- dcm_ml_result$data
hfref_data <- hfref_ml_result$data
acs_data <- acs_ml_result$data
cad_data <- cad_ml_result$data
```

```{r helper-functions}
#| code-fold: true

# Extract study center from patient ID
extract_center <- function(pat_id) {
  sub("-[0-9]+$", "", pat_id)
}

acs_data %>% colnames()

# Analyze performance by center
analyze_center_performance <- function(data, disease_name, top_models = 3,
                                       min_cases_for_sampling = 15,
                                       n_controls_to_sample = NULL) {
  # Add center information
  data <- data %>%
    mutate(center = extract_center(pat_id))

  # Get top performing models
  pred_cols <- names(data)[grepl(paste0(".pred_", disease_name, "_"), names(data))]

  if (length(pred_cols) == 0) {
    return(list(summary = tibble(), data = data, augmented_data = data))
  }

  # Select top models based on overall AUC
  model_aucs <- sapply(pred_cols[1:min(top_models, length(pred_cols))], function(pred_col) {
    analysis_data <- data %>%
      filter(!is.na(.data[[pred_col]])) %>%
      mutate(disease_binary = ifelse(disease == disease_name, 1, 0))

    if (nrow(analysis_data) < 10) {
      return(0)
    }

    tryCatch(
      {
        as.numeric(roc(analysis_data$disease_binary,
          analysis_data[[pred_col]],
          quiet = TRUE
        )$auc)
      },
      error = function(e) 0
    )
  })

  top_pred_cols <- names(sort(model_aucs, decreasing = TRUE))[1:min(top_models, length(model_aucs))]

  # Create augmented dataset with sampled controls for centers without them
  data_augmented <- data

  # Get all available controls
  all_controls <- data %>%
    filter(disease == "control") %>%
    select(pat_id, center, disease, age, sex, starts_with("hsa_"), starts_with(".pred_"))

  # Identify centers needing controls
  center_summary <- data %>%
    mutate(disease_binary = ifelse(disease == disease_name, 1, 0)) %>%
    group_by(center) %>%
    summarise(
      n_cases = sum(disease_binary == 1),
      n_controls = sum(disease_binary == 0),
      .groups = "drop"
    )

  centers_needing_controls <- center_summary %>%
    filter(n_controls == 0, n_cases >= min_cases_for_sampling) %>%
    pull(center)

  if (length(centers_needing_controls) > 0 && nrow(all_controls) > 0) {
    cat(glue("\n=== AUGMENTING DATA FOR {toupper(disease_name)} ===\n"))
    cat(glue("Centers without controls but ≥{min_cases_for_sampling} cases: {paste(centers_needing_controls, collapse=', ')}\n"))

    # Sample controls for each center
    sampled_controls_list <- list()

    for (center_name in centers_needing_controls) {
      n_cases_center <- center_summary %>%
        filter(center == center_name) %>%
        pull(n_cases)

      # Determine number of controls to sample (default: match number of cases)
      n_to_sample <- ifelse(is.null(n_controls_to_sample), n_cases_center, n_controls_to_sample)
      n_to_sample <- min(n_to_sample, nrow(all_controls))

      # Sample controls from other centers
      set.seed(42) # For reproducibility
      sampled <- all_controls %>%
        slice_sample(n = n_to_sample) %>%
        mutate(
          center_original = center,
          center = center_name,
          sampled_control = TRUE
        )

      sampled_controls_list[[center_name]] <- sampled

      cat(glue("  {center_name}: sampled {n_to_sample} controls (from {nrow(all_controls)} available)\n"))
    }

    # Combine sampled controls
    sampled_controls <- bind_rows(sampled_controls_list)

    # Add to augmented data
    data_augmented <- data %>%
      mutate(sampled_control = FALSE) %>%
      bind_rows(sampled_controls)

    cat(glue("Total controls sampled: {nrow(sampled_controls)}\n\n"))
  }

  # Analyze by center (using augmented data)
  results <- tibble()

  for (pred_col in top_pred_cols) {
    # Store the full model identifier (everything after ".pred_")
    model_name <- str_replace(pred_col, "^\\.pred_", "")

    center_results <- data_augmented %>%
      filter(!is.na(.data[[pred_col]])) %>%
      mutate(disease_binary = ifelse(disease == disease_name, 1, 0)) %>%
      group_by(center) %>%
      summarise(
        n_total = n(),
        n_cases = sum(disease_binary == 1),
        n_controls = sum(disease_binary == 0),
        n_sampled_controls = sum(sampled_control == TRUE, na.rm = TRUE),
        auc = tryCatch(
          {
            if (n() >= 10 && length(unique(disease_binary)) == 2) {
              as.numeric(roc(disease_binary, .data[[pred_col]], quiet = TRUE)$auc)
            } else {
              NA_real_
            }
          },
          error = function(e) NA_real_
        ),
        .groups = "drop"
      ) %>%
      mutate(
        model = model_name,
        disease = disease_name
      )

    results <- bind_rows(results, center_results)
  }

  return(list(summary = results, data = data, augmented_data = data_augmented))
}

# Plot center performance
plot_center_performance <- function(center_results, disease_name) {
  # Filter to centers with valid AUC
  plot_data <- center_results %>%
    filter(!is.na(auc), n_total >= 10) %>%
    mutate(
      model_clean = str_replace_all(model, "_", " "),
      center_label = glue("{center}\n(n={n_total})"),
      has_sampled = n_sampled_controls > 0
    )

  if (nrow(plot_data) == 0) {
    return(NULL)
  }

  # Build a dynamic color palette for the number of models to plot
  model_levels <- unique(plot_data$model_clean)
  base_palette <- c(
    "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd",
    "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf"
  )
  fill_values <- setNames(rep(base_palette, length.out = length(model_levels)), model_levels)

  # AUC by center plot
  p1 <- ggplot(plot_data, aes(x = reorder(center_label, -auc), y = auc, fill = model_clean)) +
    geom_col(position = position_dodge(width = 0.8), alpha = 0.8) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = 0.7, linetype = "dotted", color = "gray60") +
    # Add marker for centers with sampled controls
    geom_point(
      data = plot_data %>% filter(has_sampled),
      aes(x = center_label, y = 1.05),
      shape = 8, size = 3, color = "#d62728", inherit.aes = FALSE
    ) +
    scale_y_continuous(limits = c(0, 1.1), breaks = seq(0, 1, 0.1)) +
    scale_fill_manual(values = fill_values) +
    labs(
      title = glue("{toupper(disease_name)}: ML Model Performance by Study Center"),
      subtitle = "* indicates centers with sampled controls",
      x = "Study Center",
      y = "AUC",
      fill = "ML Model"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  # Sample size plot with distinction for sampled controls
  p2 <- plot_data %>%
    distinct(center, n_total, n_cases, n_controls, n_sampled_controls) %>%
    mutate(
      n_controls_original = n_controls - n_sampled_controls,
      center_label = glue("{center}\n(n={n_total})")
    ) %>%
    select(center, center_label, n_cases, n_controls_original, n_sampled_controls) %>%
    pivot_longer(
      cols = c(n_cases, n_controls_original, n_sampled_controls),
      names_to = "group", values_to = "count"
    ) %>%
    filter(count > 0) %>%
    mutate(
      group = case_when(
        group == "n_cases" ~ disease_name,
        group == "n_controls_original" ~ "control (original)",
        group == "n_sampled_controls" ~ "control (sampled)"
      ),
      group = factor(group, levels = c(disease_name, "control (original)", "control (sampled)"))
    ) %>%
    ggplot(aes(x = reorder(center_label, -count), y = count, fill = group)) +
    geom_col(alpha = 0.8) +
    scale_fill_manual(values = c(
      setNames("#1f77b4", disease_name),
      "control (original)" = "#2ca02c",
      "control (sampled)" = "#ff7f0e"
    )) +
    labs(
      title = "Sample Distribution by Center",
      subtitle = "Orange indicates sampled controls from other centers",
      x = "Study Center",
      y = "Number of Patients",
      fill = "Group"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank()
    )

  return(list(auc_plot = p1, sample_plot = p2))
}

# Plot ROC curves by center
plot_roc_by_center <- function(data, disease_name, center_summary = NULL,
                               show_model = 1, min_samples = 5) {
  # If center_summary provided, use it to select the model by rank
  # Otherwise, expect model_name to be passed (backwards compatibility)
  if (!is.null(center_summary)) {
    # Get the nth best model based on mean AUC across centers
    model_ranking <- center_summary %>%
      filter(!is.na(auc)) %>%
      group_by(model) %>%
      summarise(mean_auc = mean(auc, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(mean_auc))

    if (show_model > nrow(model_ranking)) {
      cat(sprintf(
        "Requested model rank %d but only %d models available\n",
        show_model, nrow(model_ranking)
      ))
      return(NULL)
    }

    model_name <- model_ranking$model[show_model]
    model_rank_auc <- model_ranking$mean_auc[show_model]

    cat(sprintf(
      "Showing ROC curves for model rank #%d: %s (mean AUC=%.3f)\n",
      show_model, model_name, model_rank_auc
    ))
  } else {
    # Backwards compatibility - should not reach here with new code
    stop("center_summary must be provided")
  }

  # Extract center and prepare data
  # The augmented data already has the center column with correct assignments
  analysis_data <- data %>%
    mutate(
      disease_binary = ifelse(disease == disease_name, 1, 0)
    )

  # Check if we have sampled_control column (indicating augmented data)
  has_sampled_controls <- "sampled_control" %in% names(analysis_data)
  cat(sprintf(
    "Using %s data (has sampled controls: %s)\n",
    ifelse(has_sampled_controls, "augmented", "original"), has_sampled_controls
  ))

  # Get the prediction column for this model
  # model_name already contains the full suffix (e.g., "dcm_glmnet_dcm")
  pred_col <- paste0(".pred_", model_name)

  if (!pred_col %in% names(analysis_data)) {
    cat("Prediction column not found:", pred_col, "\n")
    cat("Available prediction columns:", paste(grep("\\.pred_", names(analysis_data), value = TRUE), collapse = ", "), "\n")
    return(NULL)
  }

  # Filter to valid data
  analysis_data <- analysis_data %>%
    filter(!is.na(.data[[pred_col]]))

  # Calculate ROC curves for each center
  roc_data_list <- list()
  center_aucs <- tibble()

  # Show all centers first, then filter
  all_centers_summary <- analysis_data %>%
    group_by(center) %>%
    summarise(
      n_total = n(),
      n_cases = sum(disease_binary == 1),
      n_controls = sum(disease_binary == 0),
      n_sampled = ifelse(has_sampled_controls, sum(sampled_control == TRUE, na.rm = TRUE), 0),
      .groups = "drop"
    )

  cat("All centers summary:\n")
  for (i in 1:nrow(all_centers_summary)) {
    row <- all_centers_summary[i, ]
    cat(sprintf(
      "  %s: total=%d, cases=%d, controls=%d, sampled=%d\n",
      row$center, row$n_total, row$n_cases, row$n_controls, row$n_sampled
    ))
  }

  centers <- analysis_data %>%
    count(center) %>%
    filter(n >= min_samples) %>%
    pull(center)

  cat(sprintf(
    "\nCenters with ≥%d samples: %d (%s)\n",
    min_samples, length(centers), paste(centers, collapse = ", ")
  ))

  for (center_name in centers) {
    center_data <- analysis_data %>%
      filter(center == center_name)

    # Check if we have both classes
    n_classes <- length(unique(center_data$disease_binary))
    n_cases <- sum(center_data$disease_binary == 1)
    n_controls <- sum(center_data$disease_binary == 0)
    n_sampled <- ifelse(has_sampled_controls, sum(center_data$sampled_control == TRUE, na.rm = TRUE), 0)

    cat(sprintf(
      "  %s: n=%d, cases=%d, controls=%d (sampled=%d), classes=%d\n",
      center_name, nrow(center_data), n_cases, n_controls, n_sampled, n_classes
    ))

    if (n_classes < 2) {
      cat(sprintf("    Skipping %s - insufficient classes\n", center_name))
      next
    }

    # Calculate ROC
    roc_obj <- tryCatch(
      {
        roc(center_data$disease_binary,
          center_data[[pred_col]],
          quiet = TRUE
        )
      },
      error = function(e) NULL
    )

    if (!is.null(roc_obj)) {
      auc_val <- as.numeric(roc_obj$auc)
      cat(sprintf("    ROC calculated successfully - AUC=%.3f\n", auc_val))

      # Store ROC curve data
      roc_data_list[[center_name]] <- tibble(
        center = center_name,
        specificity = roc_obj$specificities,
        sensitivity = roc_obj$sensitivities,
        auc = auc_val,
        n = nrow(center_data)
      )

      # Store AUC summary
      center_aucs <- bind_rows(
        center_aucs,
        tibble(center = center_name, auc = auc_val, n = nrow(center_data))
      )
    } else {
      cat(sprintf("    ROC calculation failed for %s\n", center_name))
    }
  }

  if (length(roc_data_list) == 0) {
    cat("No ROC curves could be calculated - returning NULL\n")
    return(NULL)
  }

  cat(sprintf("Successfully calculated ROC curves for %d centers\n", length(roc_data_list)))

  # Combine all ROC data
  roc_data <- bind_rows(roc_data_list) %>%
    mutate(center_label = glue("{center} (AUC={round(auc, 3)}, n={n})"))

  # Create ROC plot
  model_clean <- str_replace_all(model_name, "_", " ")
  rank_label <- ifelse(!is.null(center_summary),
    sprintf("Model rank #%d", show_model),
    ""
  )

  # Dynamic color palette for centers (supports any number of centers)
  center_levels <- unique(roc_data$center_label)
  base_palette <- c(
    "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd",
    "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf"
  )
  color_values <- setNames(rep(base_palette, length.out = length(center_levels)), center_levels)

  p <- ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = center_label)) +
    geom_line(size = 1.2, alpha = 0.8) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    scale_color_manual(name = "Study Center", values = color_values) +
    coord_equal() +
    labs(
      title = glue("{toupper(disease_name)}: ROC Curves by Study Center"),
      subtitle = glue("{rank_label} - Model: {model_clean}"),
      x = "1 - Specificity (False Positive Rate)",
      y = "Sensitivity (True Positive Rate)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "right",
      legend.text = element_text(size = 10),
      panel.grid.minor = element_blank()
    ) +
    guides(color = guide_legend(ncol = 1))

  return(p)
}
```

# Analysis

## DCM Performance by Center

```{r dcm-analysis}
#| echo: false
#| warning: false

# Set minimum cases threshold for control sampling
MIN_CASES_THRESHOLD <- 15

dcm_center <- analyze_center_performance(
  dcm_data,
  "dcm",
  top_models = 3,
  min_cases_for_sampling = MIN_CASES_THRESHOLD
)

if (nrow(dcm_center$summary) > 0) {
  # Summary table
  dcm_center$summary %>%
    filter(!is.na(auc)) %>%
    mutate(
      model = str_replace_all(model, "_", " "),
      auc = round(auc, 3),
      center_note = ifelse(n_sampled_controls > 0,
        paste0(center, " *"),
        center
      )
    ) %>%
    arrange(center, desc(auc)) %>%
    select(center_note, model, n_total, n_cases, n_controls, n_sampled_controls, auc) %>%
    kable(
      col.names = c("Center", "ML Model", "Total N", "Cases", "Controls", "Sampled", "AUC"),
      caption = "DCM: ML Model Performance by Study Center"
    ) %>%
    kable_styling(bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE)

  # Plots
  dcm_plots <- plot_center_performance(dcm_center$summary, "dcm")
  if (!is.null(dcm_plots)) {
    # Save plots as SVG
    ggsave("scripts2025/figures/ml_center_performance/dcm_auc_by_center.svg",
      dcm_plots$auc_plot,
      width = 8, height = 6
    )
    ggsave("scripts2025/figures/ml_center_performance/dcm_sample_distribution.svg",
      dcm_plots$sample_plot,
      width = 8, height = 6
    )

    print(dcm_plots$auc_plot)
    print(dcm_plots$sample_plot)
  }

  # ROC curves by center
  # Change show_model parameter to select different model ranks:
  # show_model = 1 (best model), show_model = 2 (second best), etc.
  # NOTE: Using augmented data to include centers with sampled controls
  dcm_roc_plot <- plot_roc_by_center(
    dcm_center$augmented_data, # Use augmented data with sampled controls
    "dcm",
    center_summary = dcm_center$summary,
    show_model = 1 # Show best performing model
  )

  if (!is.null(dcm_roc_plot)) {
    # Save ROC plot as SVG
    ggsave("scripts2025/figures/ml_center_performance/dcm_roc_by_center_model1.svg",
      dcm_roc_plot,
      width = 6, height = 6
    )

    print(dcm_roc_plot)
  }

  # Save results
  write_csv(dcm_center$summary, "output/dcm_center_performance.csv")
}
```

## HFrEF Performance by Center

```{r hfref-analysis}
#| echo: false
#| warning: false

hfref_center <- analyze_center_performance(
  hfref_data,
  "hfref",
  top_models = 9,
  min_cases_for_sampling = MIN_CASES_THRESHOLD
)

if (nrow(hfref_center$summary) > 0) {
  # Summary table
  hfref_center$summary %>%
    filter(!is.na(auc)) %>%
    mutate(
      model = str_replace_all(model, "_", " "),
      auc = round(auc, 3),
      center_note = ifelse(n_sampled_controls > 0,
        paste0(center, " *"),
        center
      )
    ) %>%
    arrange(center, desc(auc)) %>%
    select(center_note, model, n_total, n_cases, n_controls, n_sampled_controls, auc) %>%
    kable(
      col.names = c("Center", "ML Model", "Total N", "Cases", "Controls", "Sampled", "AUC"),
      caption = "HFrEF: ML Model Performance by Study Center"
    ) %>%
    kable_styling(bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE)

  # Plots
  hfref_plots <- plot_center_performance(hfref_center$summary, "hfref")
  if (!is.null(hfref_plots)) {
    # Save plots as SVG
    ggsave("scripts2025/figures/ml_center_performance/hfref_auc_by_center.svg",
      hfref_plots$auc_plot,
      width = 8, height = 6
    )
    ggsave("scripts2025/figures/ml_center_performance/hfref_sample_distribution.svg",
      hfref_plots$sample_plot,
      width = 8, height = 6
    )

    print(hfref_plots$auc_plot)
    print(hfref_plots$sample_plot)
  }

  # ROC curves by center
  # Change show_model parameter to select different model ranks:
  # show_model = 1 (best model), show_model = 2 (second best), etc.
  # NOTE: Using augmented data to include centers with sampled controls
  hfref_roc_plot <- plot_roc_by_center(
    hfref_center$augmented_data, # Use augmented data with sampled controls
    "hfref",
    center_summary = hfref_center$summary,
    show_model = 8 # Show best performing model
  )

  if (!is.null(hfref_roc_plot)) {
    # Save ROC plot as SVG
    ggsave("scripts2025/figures/ml_center_performance/hfref_roc_by_center_model8.svg",
      hfref_roc_plot,
      width = 6, height = 6
    )

    print(hfref_roc_plot)
  }

  # Save results
  write_csv(hfref_center$summary, "output/hfref_center_performance.csv")
}
```

## ACS Performance by Center

```{r acs-analysis}
#| echo: false
#| warning: false

acs_center <- analyze_center_performance(
  acs_data,
  "acs",
  top_models = 3,
  min_cases_for_sampling = MIN_CASES_THRESHOLD
)

if (nrow(acs_center$summary) > 0) {
  # Summary table
  acs_center$summary %>%
    filter(!is.na(auc)) %>%
    mutate(
      model = str_replace_all(model, "_", " "),
      auc = round(auc, 3),
      center_note = ifelse(n_sampled_controls > 0,
        paste0(center, " *"),
        center
      )
    ) %>%
    arrange(center, desc(auc)) %>%
    select(center_note, model, n_total, n_cases, n_controls, n_sampled_controls, auc) %>%
    kable(
      col.names = c("Center", "ML Model", "Total N", "Cases", "Controls", "Sampled", "AUC"),
      caption = "ACS: ML Model Performance by Study Center"
    ) %>%
    kable_styling(bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE)

  # Plots
  acs_plots <- plot_center_performance(acs_center$summary, "acs")
  if (!is.null(acs_plots)) {
    # Save plots as SVG
    ggsave("scripts2025/figures/ml_center_performance/acs_auc_by_center.svg",
      acs_plots$auc_plot,
      width = 8, height = 6
    )
    ggsave("scripts2025/figures/ml_center_performance/acs_sample_distribution.svg",
      acs_plots$sample_plot,
      width = 8, height = 6
    )

    print(acs_plots$auc_plot)
    print(acs_plots$sample_plot)
  }

  # ROC curves by center
  # Change show_model parameter to select different model ranks:
  # show_model = 1 (best model), show_model = 2 (second best), etc.
  # NOTE: Using augmented data to include centers with sampled controls
  acs_roc_plot <- plot_roc_by_center(
    acs_center$augmented_data, # Use augmented data with sampled controls
    "acs",
    center_summary = acs_center$summary,
    show_model = 1 # Show best performing model
  )

  if (!is.null(acs_roc_plot)) {
    # Save ROC plot as SVG
    ggsave("scripts2025/figures/ml_center_performance/acs_roc_by_center_model1.svg",
      acs_roc_plot,
      width = 6, height = 6
    )

    print(acs_roc_plot)
  }

  # Save results
  write_csv(acs_center$summary, "output/acs_center_performance.csv")
}
```

## CAD Performance by Center

```{r cad-analysis}
#| echo: false
#| warning: false

cad_center <- analyze_center_performance(
  cad_data,
  "cad",
  top_models = 3,
  min_cases_for_sampling = MIN_CASES_THRESHOLD
)

if (nrow(cad_center$summary) > 0) {
  # Summary table
  cad_center$summary %>%
    filter(!is.na(auc)) %>%
    mutate(
      model = str_replace_all(model, "_", " "),
      auc = round(auc, 3),
      center_note = ifelse(n_sampled_controls > 0,
        paste0(center, " *"),
        center
      )
    ) %>%
    arrange(center, desc(auc)) %>%
    select(center_note, model, n_total, n_cases, n_controls, n_sampled_controls, auc) %>%
    kable(
      col.names = c("Center", "ML Model", "Total N", "Cases", "Controls", "Sampled", "AUC"),
      caption = "CAD: ML Model Performance by Study Center"
    ) %>%
    kable_styling(bootstrap_options = c("striped", "hover")) %>%
    column_spec(1, bold = TRUE)

  # Plots
  cad_plots <- plot_center_performance(cad_center$summary, "cad")
  if (!is.null(cad_plots)) {
    # Save plots as SVG
    ggsave("scripts2025/figures/ml_center_performance/cad_auc_by_center.svg",
      cad_plots$auc_plot,
      width = 8, height = 6
    )
    ggsave("scripts2025/figures/ml_center_performance/cad_sample_distribution.svg",
      cad_plots$sample_plot,
      width = 8, height = 6
    )

    print(cad_plots$auc_plot)
    print(cad_plots$sample_plot)
  }

  # ROC curves by center
  # Change show_model parameter to select different model ranks:
  # show_model = 1 (best model), show_model = 2 (second best), etc.
  # NOTE: Using augmented data to include centers with sampled controls
  cad_roc_plot <- plot_roc_by_center(
    cad_center$augmented_data, # Use augmented data with sampled controls
    "cad",
    center_summary = cad_center$summary,
    show_model = 3 # Show best performing model
  )

  if (!is.null(cad_roc_plot)) {
    # Save ROC plot as SVG
    ggsave("scripts2025/figures/ml_center_performance/cad_roc_by_center_model3.svg",
      cad_roc_plot,
      width = 6, height = 6
    )

    print(cad_roc_plot)
  }

  # Save results
  write_csv(cad_center$summary, "output/cad_center_performance.csv")
}
```

# Summary

This analysis examines ML model performance across different study centers in the BestAgeing consortium. 

## Methodology

**Control Sampling Strategy:**
For centers that enrolled ≥`r MIN_CASES_THRESHOLD` disease cases but no controls, we randomly sampled controls from other centers to enable performance evaluation. This approach allows assessment of model generalizability across centers while acknowledging the study design where some centers focused exclusively on disease recruitment.

- **Threshold**: Centers with ≥`r MIN_CASES_THRESHOLD` cases qualify for control sampling
- **Sampling**: Controls randomly selected from all available controls (with replacement across centers)
- **Matching**: Number of sampled controls matches the number of cases in that center
- **Reproducibility**: Fixed seed (42) ensures consistent results

## Key Findings

- **Center Variability**: Performance varies across centers, reflecting differences in patient populations and sample sizes
- **Sample Size Impact**: Centers with larger sample sizes generally show more stable performance metrics
- **Model Consistency**: Top-performing ML models maintain competitive AUC across most centers
- **Sampled Controls**: Centers marked with * use randomly sampled controls for evaluation
- **Disease-Specific Patterns**: All four cardiovascular diseases (DCM, HFrEF, ACS, CAD) evaluated across centers

**Interpretation Notes:**
- Centers with <10 patients excluded from analysis
- AUC values should be interpreted in context of sample size
- Performance differences may reflect population heterogeneity or batch effects

---

*Analysis conducted on `r Sys.Date()`. Results saved to output/ directory.*
